\documentclass{tex/vcbook-float}
\usepackage{tex/poemtranslation}
\begin{document}
% Ha_de_los_coros.tex
\begin{poemtranslation}
    \begin{original}
        \Stanza{4}
        1. (i.) Ha de los Coros del cielo? &
        cuyo misterioso canto &
        sagrada canción entona &
        por los siglos de los Astros. \&

        \Stanza{4}
        2. (ii.) Quién interrompe lo acorde &
        deste sagrado Palacio? &
        que la celeste armonía &
        pendiente está de su labio. \&

        \Stanza{4}
        3. (i.) Una novedad suspenda &
        vuestros Músicos sagrados, &
        pues suspende a las esferas &
        lo voluble de sus arcos. \&

        \Stanza{4}
        4. (ii.) Hagan pausa los zafiros &
        deste cuaderno estrellado, &
        dejen doblada la hoja &
        los azules cartapacios. \&

        \Stanza{4}
        5. (i.) La consonancia más nueva &
        escuchad, que la han formado, &
        hombre, y Dios en un pesebre, &
        cielo, y tierra en un establo; \&

        \Stanza{4}
        6. (ii.)  Despuéblense las esferas &
        de celestes ciudadanos, &
        pues bajan a ser sus glorias &
        alegrías de los campos. \&
    \end{original}

    \begin{translation}
        \Stanza{4}
        1. Ah, the choirs of heaven! &
        whose mysterious chant &
        intones a sacred song &
        upon the centuries of the stars. \&

        \Stanza{4}
        2. Who interrupts the concord &
        of this sacred palace? &
        for the heavenly harmony &
        is hanging on his lips. \&
 
        \Stanza{4}
        3. A novelty suspends &
        your sacred musicians, &
        just as it suspends for the stars &
        the fickleness of their curving motions. \&

        \Stanza{4}
        4. The sapphires of this starry note-book &
        make a rest, &
        the azure partbooks &
        let the folded sheet fall closed. \&

        \Stanza{4}
        5. Listen to the newest consonance, &
        for they have formed it--- &
        man and God in a manger; &
        heaven and earth in a stable. \&

        \Stanza{4}
        6. Let the spheres be emptied &
        of heavenly citizens, &
        for they descend to be the merry &
        \term{Glorias} of the fields. \&
    \end{translation}
\end{poemtranslation}
\end{document}


