\begin{poemtranslation}
    \begin{original}
        \Stanza{4}
        1. Las fugas que el primer hombre &
        formó en desatentos pasos &
        al compás ajusta un Niño &
        de las perlas de su llanto. \&

        \Stanza{4}
        2. Qué mucho si a los despeños &
        que le ocasionó un engaño, &
        bella corriente de aljófar, &
        grillos le previene blandos \&

        \Stanza{4}
        3. Una voz que ha dado el cielo, &
        de metal más soberano &
        a ordenar entra sonora &
        la disonancia del barro. \&

        \Stanza{4}
        4. Concierto tan soberano &
        sólo pudo ser reparo, &
        con una voz tan humilde, &
        de un desentono tan vano \&

        \Stanza{4}
        5. En las pajas sustenido &
        dulcemente se ha escuchado &
        ligar en pajas lo eterno, &
        reducir lo inmenso a espacio. \&

        \Stanza{4}
        6. Divina cláusula sea &
        deste eterno canto llano, &
        que forma en su movimiento &
        de cada punto un milagro. \&
    \end{original}

    \begin{translation}
        \Stanza{4}
        1. The flight/fugue that the first man &
        made in heedless paces &
        is set aright by a baby boy to the measure &
        of the pearls of his crying. \&

        \Stanza{4}
        2. What wonder, if from the falls &
        that a deceit caused him, &
        the lovely mother-of-pearl stream &
        gently restrains him with shackles. \&

        \Stanza{4}
        3. A voice that heaven has given, &
        of the most sovereign timbre, &
        to bring order, enters resounding &
        into the dissonance of the clay. \&

        \Stanza{4}
        4. So sovereign a concord/concerto & 
        could only be a resolution, &
        with so humble a voice, &
        of so vain a discord. \&

        \Stanza{4}
        5. Upon the straw sustained &
        sweetly he has been heard &
        binding in straw the eternal, &
        reducing the immense to this space/slowly. \&

        \Stanza{4}
        6. Let there be a divine cadence &
        of this eternal plainsong, &
        which forms in its movement &
        a miracle from each note/point. \&
    \end{translation}
\end{poemtranslation}
\endinput
