\begin{poemtranslation}
    \begin{original}
    \StanzaSection{4}[\add{Introducción}]
        Cantores \uline{de la Capilla,} &
        \uline{cuenta con lo que se canta,} &
        \uline{que es Músico el} Niño, \uline{y nota} &
        \uline{las mas leves dissonancias.} \&

        \Stanza{4}
        La música que \uline{componó} &
        de vozes \uline{altas} y bajas, &
        a compás mayor las rige, &
        y es proporción abreviada. \&

        \Stanza{4}
        Una \uline{clave} con \uline{tres} tiempos &
        \uline{pone} con \uline{destreza} tanta, &
        que el pasado y el futuro &
        al \uline{compás} presente iguala. \&

        \Stanza{4}
        Un Coro errado enmendó &
        con un \uline{medio}, que a la entrada &
        puso, y una espiración &
        que \uline{para el} Calvario \uline{guarda.} 
        \SectionBreak

        \StanzaSection{16}[Estribo]
        O que lindamente suenan! &
        o que dulcemente cantan &
        al compás que lleva \uline{el Infante,} &
        Serafines que cruzan y passan! &
        y de sus gemidos &
        aprended trinados y sustenidos, &
        y con mil primores &
        responden los Reyes y los Pastores, &
        después que \uline{aguardaron uno} &
        que llegó a \uline{tiempo oportuno}, &
        \uline{quien antes del tiempo fue} &
        \uline{por el signo a la mi re,} &
        \uline{puestos los ojos en mi,} &
        con que mil maravillas vi &
        \uline{tan de ir} \add{sic} \uline{y de admirar,} &
        que si lo acierto a dezir, \&

        \Stanza{2}
        \uline{todo en el hombre es subir,} &
        \uline{y todo en Dios es bajar.}
        \SectionBreak

        \StanzaSection{4}[Coplas]
        1. A Suspensiones el Cielo &
        hace en sus esferas pausas, &
        porque el Ángel que los mueve &
        cuanto le admira le pasma. \&

        \Stanza{4}
        2. Los compases son del tiempo, &
        que ya sus compases guarda &
        quien al tiempo lleva \uline{siglos,} &
        quien lleva al siglo \uline{distancias.} \&

        \Stanza{4}
        3. Tambien se canta a ternario, & 
        pues entran, caben y passan & 
        tres Reyes en un compas, & 
        de \uline{brutos} \uline{breve} morada. \& 

        \Stanza{4}
        4. \uline{Por sol comienza una gloria,} & 
        \uline{por mi se canta una gracia,} & 
        \uline{y a medio compás la Noche} & 
        \uline{remeda quiebros del Alba.} \&
    \end{original}

    \begin{translation}
        \StanzaSection{4}
        Singers of the chapel choir, &
        keep count with what is sung, &
        for the Child is a musician, and he notes &
        even the most venial dissonances. \&

        \Stanza{4}
        The music that he composed &
        of voices high and low, &
        directs \add{the voices} \term{alla breve}, &
        and it is a reduced proportion. \&

        \Stanza{4}
        A signature of three times \add{meters} &
        he places with such dexterity &
        that he makes the past and the present &
        equal the present measure. \&

        \Stanza{4}
        He amended an errant choir &
        through a mean \add{rest?} that he placed &
        at the entrance, and a breath \add{short rest} &
        that he preserves for Calvary.
        \SectionBreak

        \StanzaSection{16}
        O, how lovely they sound! &
        O how sweetly they sing &
        to the measure that the Infant gives--- &
        seraphim that cross and pass! &
        And from his sighs &
        learn trills and sharps, &
        and with a thousand excellences &
        the Kings and shepherds respond, &
        after they waited for one &
        who arrived at the opportune time, &
        who was before time, &
        upon the sign of \term{A (la, mi, re)} &
        his eyes set on me \add{\term{mi}}, &
        with what thousand wonders I saw, &
        as much to go \add{\term{recte:} hear} and to admire, &
        for if I say it precisely: \&

        \Stanza{2}
        Everything in Man is to ascend, &
        and everything in God is to descend. 
        \SectionBreak

        \StanzaSection{4}
        1. In suspensions, Heaven &
        makes rests in its spheres, &
        because the angel that moves them, &
        the more he admires, the more he is stunned. \&

        \Stanza{4}
        2. The measures are of time, &
        for he still guards his measures &
        the one who at one time carries centuries, &
        who to this century bears the intervals. \&

        \Stanza{4}
        3. It is also sung in triple time, &
        for they enter, fit, and pass, &
        three Kings in one measure, &
        a brief rest for the animals. \&

        \Stanza{4}
        4. With the sun \add{on \term{sol}} a \quoted{glory} begins, &
        upon me \add{\term{mi}} a \quoted{grace} is sung, &
        and at the half-measure, the night &
        imitates the trills of the dawn. \&

    \end{translation}
\end{poemtranslation}

