% vim: set foldmethod=marker :

% Cashner, *Hearing Faith*
% chapter 1: Villancicos as Musical Theology
% 
% 2015-03-18	Dissertation defended
% 2017-11-15    New start for book proposal
% 2018-05-21    Converted back to LaTeX
% 2018-07-25    Expanded for press readers
% 2019-07-08	New start for Brill, under contract
% 2019-07-21    Complete revised draft text
% 2019-08-09    Begin revision after ch1-5 draft complete

\part{Listening for Faith}
\label{part:faith}

\chapter{Villancicos as Musical Theology}
\label{ch:intro}

\epigraph
{ergo fides ex auditu\\
auditus autem per verbum Christi}
{Romans 10:17}

%{{{1 intro
%{{{2 intro^2
\quoted{Faith comes through hearing}, wrote Paul the apostle to the Christian
community in first-century Rome, \quoted{and what is heard, by the word of
Christ} (\scripture{Rom}{10:17}).
Sixteen centuries later, amid the ongoing reformations of the Western Church,
Christians were finding ever new ways to make faith audible.
Voices raised in acrid contention or pious devotion boomed from pulpits,
clamored in public squares, and were echoed in homes and schools.  
In new forms of vernacular music, the voices of the newly distinct communities
united to articulate their own vision of Christian faith.
Roman Catholic reformers and missionaries, charged by the Council of
Trent (\XXX[years]) to educate and evangelize, enlisted music in their
campaigns to build Christian civilization, both in an increasingly divided
Europe and in the expanding global domains of the Spanish crown.
In these efforts to make \quoted{the word of Christ} to be heard and
believed---to make faith appeal to hearing---what did they understand to be the
role of music?
What kind of power did Catholics believe music held to affect the relationship
between hearing and faith?

This book is a study of how Christians in early modern Spain and Spanish
America enacted religious beliefs about music, through the medium of music
itself.
It focuses on a genre of devotional music known as \term{villancicos}, which
were musical settings of poetry in vernacular language, most often Castilian
Spanish.
A genre that had previously been an elite form of courtly entertainment and
sometimes devotion was transformed around the turn of the seventeenth century
into variety of complex, large-scale forms of vocal and instrumental music
performed as an integral part of public church rituals.

Of all the musical forms of Catholic Spain, sacred villancicos address the
theological nature and function of music most frequently and directly.
Hundreds of surviving poems and music begin with calls to
listen---\foreign{Escuchad}, \foreign{Oíd}, foreign{Atended}---and a large
proportion of villancicos explicitly refer to music in their texts, in some
cases using musical techniques and terms as metaphors for theological concepts.
One of the most common occasions for villancico performances was in the liturgy
of Matins on Christmas Eve, when large ensembles of singers and musicians
celebrated the paradox of the incarnation---the most high God being born
as a helpless infant in a humble stable---with musical representations of
angel choruses and shepherds singing and dancing shepherds.
In these depictions of heavenly and earthly music, the ensembles were making
music about music.
Composers used a variety of techniques to refer to music beyond that being
performed in the moment, including evocations of birdsong, musical instruments,
quotations of specific songs and dances, and more conceptual plays on
music-theory terms, such as writing a strict fugue on the words
\mentioned{celestial counterpoint}.
If a play within a play in seventeenth-century Spanish or English theater is
metatheatrical, then these pieces are \term{metamusical}.

Through this genre of musical performance people embodied their theological
conceptions of music through the structures of music itself.
I call this \term{musical theology}, by which I mean a practice that goes
beyond the theological discourse about music found in written treatises or the
use of music for religious purposes, and becomes a way of \emph{performing}
theology.
This is a form of music that embodies the beliefs it proclaims.
Charles Seeger theorized that music and language were two distinct forms of
discourse and ways of knowing, so that just as we can speak about music by
using verbal discourse to refer to musical discourse, we could also
\emph{music} about music.
    \citXXX[Seeger:Unitary, Small:Musicking]
I argue that metamusical villancicos were precisely this kind of musical
discourse about music.
%}}}2

%{{{2 singing about singing : examples
\subsection{Singing about Singing}

Let us begin by listening to two villancicos that embody \quoted{singing about
singing}, while serving as representative examples of the genre.%
    \autocites{Murata:Singing}
    [\XXX]{Illari:Polychoral}
The first piece was composed by Juan Gutiérrez de Padilla (\circa{1590}--1664)
for the Cathedral of Puebla de los Ángeles, the most important religious center
in Spanish America.%
\begin{Footnote}
    \sig{MEX-PC}{Leg. 1/2}. 
    \XXX[Audio recordings of every numbered example of music and poetry in this
    book are available on the companion website.]
    Another villancico by this composer is the subject of
    \cref{ch:padilla-voces}.
\end{Footnote}
\term{En la gloria de un portalillo} was first performed in the Matins liturgy
on Christmas Eve 1652.  
In just the first seven lines of the anonymous poem
(\cref{poem:En_la_gloria_de_un_portalillo-Padilla-1652}), the villancico refers
to multiple kinds of music, referring to the sounds of voices, choirs singing,
birdsong, dancing, and using solmization syllables.

%{{5 poem GdP En la gloria
\insertPoem{En_la_gloria_de_un_portalillo-Padilla-1652}
{\wtitle{En la gloria de un portalillo}, estribillo as set by Juan Gutiérrez
de Padilla, Puebla Cathedral, Christmas 1652 
(\sig{MEX-Pc}{Leg. 1/2})}
%}}}5

Gutiérrez de Padilla's setting is metamusical in that it enacts these
references through music (\cref{mux:Padilla-Portalillo}).
It also has the virtue of demonstrating several typical features of the genre
which should provide a useful foundation for our study.
The solo line is followed by a passage of dialogue between two four-voice
choirs, concluding with an emphatic cadence for the full chorus, was typical of
polychoral technique.
The setting is in a lively triple meter (\meterCZ{} or CZ, equivalent to
\meterCThree) featuring regular rhythmic groupings of three minims (half
notes).
\begin{Footnote}
    Contemporary music writer Andrés Lorente \meterCZ{} is a cursive shorthand
    for \meterCThreeTwo or \meterCThree, all representing \term{tiempo
    imperfecto menor de proporción menor}.
    The three minims in every measure (\term{compás} in Spanish, same as
    \term{tactus}) were understood to be in a $3:2$ relationship to the two
    minims in a measure of \term{C} time (\term{tiempo imperfecto menor}).
    \citXXX[Lorente]
    See the preface/glossary/\XXX{} for more background.
\end{Footnote}
Within this meter the composer makes frequent use of \term{sesquialtera} or
hemiola, an alteration of the rhythmic pattern that feels like a momentary
shift from ternary to duple stresses.
The shifts of duple and triple stresses combine with stresses on the second
beat of the measure to create an energetic atmosphere with a rejoicing affect.  
The mutual exhortations of the choruses, with the voices of each choir declaiming
homorhythmically in the same highly rhythmic, syncopated manner as the soloist,
and with the \term{tiples} (boy sopranos) of both choirs singing at the top of
their range, would have brilliantly seized the attention of listeners.

After this introductory \term{exordium}, the Tiple I soloist continues to
describe the scene at the manger, where the shepherds \quoted{are turned to
boys}.
Playing on this expression, Gutiérrez de Padilla has the musicians
\quoted{turn} modally by adding C sharps, accented in a sesquialtera group that
contrasts with the even ternary patterns that follow.
The two-measure groups in the next passage emphasize the rhymes in
\foreign{tonos sonoros, repiten a coros} and the clear triple meter evokes the
dances of \foreign{en bailes lucidos}.
When the soloist refers to the newborn Sun, he sings the note identified in
Guidonian terminology as D \term{(la, sol, re)}---\term{sol} in the hard (G)
hexachord.  
On the same word, the bass accompanist plays a different \term{sol}, G
\term{(sol, re, ut)}.%
\begin{Footnote}
    The major Spanish music-theoretical treatises of the seventeenth century
    give full expositions of the techniques of Guidonian solmization:
    \autocites{Cerone:Melopeo}{Lorente:Porque}.
    The frequent symbolic use of Guido's syllables in villancicos suggests that
    these treatises do reflect how music was actually taught in practice.
\end{Footnote}

%{{{5 music gdp portalillo start
\insertMusic{Padilla-Portalillo}
{Gutiérrez de Padilla, \wtitle{En la gloria de un portalillo}
(\sig{MEX-Pc}{Leg. 1/2}, Christmas 1652), estribillo}
%}}}5

This villancico exemplifies music about music on several levels.
The text, which is being performed through music, itself refers to musical
performance.
The performance by the Puebla Cathedral chapel dramatizes the historical
celebration of the first Christmas while also celebrating the festival in
Padilla's present day.  
The music is self-referential on a symbolic level (as in the plays on
\term{sol}), but also functions on a more simple affective level to model and
incite affections of exuberant joy and wonder.
According to contemporary theological writers, those were the appropriate
affects for worshippers at the feast of Christmas (see
\cref{ch:padilla-voces}).

%{{{3 cererols fuera que va
Celebrating Christmas with the right kind of spirit is also the theme of
another metamusical villancico, \wtitle{Fuera, que va de invención} by Joan
Cererols (1618--1680), monk and chapelmaster at the Benedictine Abbey of
Montserrat near Barcelona.%	
\begin{Footnote}
    \sig{E-Bbc}{M/760}, \autocite[81--94]{Cererols:MEM-VC}.
    A villancico by this composer is the subject of
    \cref{ch:cererols-suspended}.
\end{Footnote}
Rather like today's catalog-like Christmas songs (\wtitle{Deck the Halls},
\wtitle{Chestnuts Roasting on an Open Fire}), the piece summons up all the
elements of a Christmas festival---masques, \foreign{zarabandas} and other
dancing, lavish decorations and clothing, pipes, drums, and so on.
As in many villancicos, the chorus acts dramatically in the role of the
festival crowd, shouting affirmations (\foreign{¡vaya!}) for each element of
the celebration as the soloists name them.  
Whereas Gutiérrez de Padilla's \wtitle{En la gloria de un portalillo} focused
primarily on the music of the historical Christmas day, the villancico of
Cererols is unambiguously about celebrating \soCalled{Christmas present}.
The piece seeks a theological meaning behind the Christmas customs: the masques
of Christmas, the poem says, are appropriate because in the Incarnation of
Christ, \foreign{Dios se disfraza} (God masks himself).
Cererols's original audience of pilgrims to the mountaintop shrine of
Montserrat would not have sung along with this piece, but the piece still
invites their wholehearted participation in the rituals of Christmas, both
through enjoying the choral singing and in the many other common-culture
customs that the piece celebrates.

The villancico allowed performers and listeners to celebrate the festival in two
senses: to sing the praises of the Christmas feast, while also singing the
praises of Christ that were appropriate to that feast. 
It did not so much teach them anything about Christ's birth as it modeled for
them appropriate modes of devotion to the Christ-child.
The primary purpose, I argue, is not to say that Christ \term{is} something but
to worship Christ \term{as} something.

These pieces presented hearers with a discourse about music, through music.
Sometimes the music they refer to is literal, human music-making; other times
it is more abstract, like the music of the spheres or the harmony of human and
divine in the incarnate Christ.
In every case, analyzing the musical choices made to represent texts about
music helps us understand how the creators and their audiences heard different
kinds of music.
And interpreting their theological aspect enables us to see how these pieces
served to communicate with hearers at a spiritual level.
%}}}3
%}}}2

%{{{2 paying attention to villancicos
\subsection{Paying Attention to Villancicos}

Villancicos continue to demand that we listen. 
Because so many villancicos explicitly address concepts of music, sensation,
and faith, these remarkable but understudied pieces offer us unique insights
into Spanish beliefs about music, which can deepen our understand of music's
role in early modern religious culture.
Villancicos constituted a major element of the soundscape of the
early modern Ibero-American world.
They engaged the creativity and piety of poets, composers, and performers
across the empire and shaped the everyday experiences of thousands of people
across social strata.
They provide evidence for a sustained effort by Spanish church leaders to use
music to make faith appeal to the sense of hearing, and they reflect widely
held beliefs and attitudes about music's spiritual power.

This book is the first major effort to understand the meanings and functions of
this music as a form of religious practice, integrating musical and theological
interpretation.
Despite their rich potential as sources for studies in several
disciplines---including literary studies, religious studies, and
musicology---few scholars have sought to understand this music in the
theological context for which it was created.
Most of the known sources of villancico music and poetry have now been
catalogued, but only a handful have been revived in performance and the
genre has received relatively little serious scholarly attention.
Even scholars of music in the Spanish Empire have overlooked these musical
sources, part of a recent trend toward a social-historical approach to music
that emphasizes critiquing colonial power and draws primarily on documentary
evidence, while in some cases deliberately excluding notated music from
consideration.%
\citXXX[baker,ramoskittrell]
On another extreme are studies that only look at musical practices and
purposefully ignore words and meaning.%
\citXXX[swadley etc]
Another study of Catholic concepts of music and hearing, focused on Italy,
provides a valuable analysis of verbal discourse \emph{about} music but does
not connect it to actual practices of music-making.
This book provides a necessary complement to these studies by analyzing how
people expressed and shaped beliefs about music through the medium of music
itself.
At the same time, the book offers a fresh approach by considering this music as
a source for historical theology, something few scholars have done.

If inquirers today wish to know what early modern Christians believed, we
must listen carefully to how they made their faith heard.
And if we wish to understand not only what music meant to early
modern people but even the details of how music worked, we must contemplate
what the makers and hearers of that music believed about its sacred power. 
In that endeavor our goal must be to understand their beliefs, not to impose
ours.
For my part I am a Christian but not a Roman Catholic, and as a child of
Enlightenment modernity and postmodernity I certainly do not share many aspects
of the early modern Catholic worldview.
But interpreting villancicos requires us to distance ourselves from our own
religious ideas---or anti-religious ideas---in order to hear the world through
historic ears, to the extent we can venture to do so.%
    \citXXX[historic ear]

By theology I do not mean the tired repetition of settled church doctrines such
as articles of the Creed or dogmas of the Council of Trent.
Instead I understand theology as a creative activity of imaginatively, even
playfully, seeking out ever-new ways of connecting revealed truth to observed
experience.
Thinking theologically in an early modern Catholic sense meant building 
endless chains of association and allusion among Biblical texts, the liturgy,
and theological writings, ancient and contemporary.
It meant interpreting new texts in light of these old ones, and reinterpeting
the old ones in light of the new.
And I will discuss below, it grew out of and reinforced a view of the world as
a book waiting to be read.

The creators of villancicos drew on common experiences of everyday life and
linked them to the sacred in inventive ways that met the spiritual needs of
specific communities.
Each piece provides a new answer to Christ's question, \quoted{With what can we
compare the kingdom of God, or what parable will we use for it?}
(\scripture{Mk}{4:21}).
The more surprising and puzzling the connection, the better---such as
representing the Virgin Mary as the chapelmaster of the heavenly chorus (see
below), or imagining Christ as a gambling card player.%
    \Autocite{Cashner:PlayingCards}

In this way villancicos embody much of Paul Ricoeur's theory of metaphor as the
bringing together of two contradictory interpretations of an utterance, in
which through a \quoted{metaphorical twist}, a novel comparison changes the
reader's view of both elements being compared.%
    \citXXX[ricoeur]
Spanish literary critics have used the term \term{conceptismo} to label the
metaphorical technique used in this kind of poetry, in which through an
extended metaphor the poem creates parallel discourses that bring two subjects
into a new relationship with each other.
As early as 196X\XXX[name], English poetry scholar \XXX[name] cited Alonso de
Ledesma's (\XXX[year]) book of devotional poetry, \wtitle{Conceptos
espirituales}, as a fountainhead for the poetic technique of villancicos that
compared worldly and ethereal, human and divine elements.%
    \citXXX[source of both and quote?]
Ledesma's earthy approach, aimed at relatively uncultivated readers, differed
starkly from the later \term{conceptista} poetry of Luis de Góngora and his
followers on both sides of the Atlantic, in which the language was pushed to
the brink of comprehensibility to make elaborate and highly learned double and
triple conceits; but both attempts at metaphorical connection were well
represented in villancico poetry.%
    \Autocites{Tenorio:Gongorismo}
    [227--228]{Gaylord:Poetry}
    {Gracian:Ingenio}
The label \term{conceptismo} alone, though, does not do much to explain how the
process of signification worked in these pieces or, more importantly, why
Spaniards enjoyed thinking this way.
It also does not help us understand the additional layer of signification added
through music, in which a poetic concept is realized through properly musical
kinds of concepts.
In religious poetry, I propose, \term{conceptismo} was a specific literary form
for theological thinking.
It provided a set of conventions through which writers used metaphor to provoke
readers to connect human and divine, temporal and eternal.
This study shows how Spanish musicians developed their own approaches in form
and style to amplify the theological thinking in the poetic texts; in other
words, they cultivated music as a form of creative theology.

The book is organized in two parts, in which the two chapters of the first part
explore the central questions about music's role in the relationship between
faith and hearing, based on a global sampling of the repertoire.
The musical and poetic sources have been edited, mostly for the first time,
from sources in nine archives in Mexico, Spain, the United Kingdom, and the
United States.
The rest of this chapter examines the different ways Spanish musicians used
the music of villancicos to refer to other kinds of music, and the theological
framework that shaped their beliefs about what this music could do.
\Cref{ch:faith-hearing} draws on villancicos on the subject of sensation and
faith, including representations of physical and spiritual deafness, to explore
the problems and anxieties Catholics encountered in their endeavors to make
faith appeal to hearing.
The three chapters of \cref{part:unhearable-music}, then, present case studies
of individual villancicos or sets of closely related pieces focused in Puebla,
Montserrat, and Zaragoza, respectively, all of which use \quoted{music about
music} to foster a listening practice in which hearers strain to move past the
simple level of audible music and rise to contemplate a higher, unhearable kind
of music.
These chapters demonstrate that Spanish composers used metamusical villancicos
to establish their place in a lineage of composition, while also revealing how
their conventions for depicting heavenly music in particular developed over the
century.
% The \XXX{glossary and the accompanying website} are meant to help fill any gaps
% in understanding for readers approaching this work from various disciplinary
% backgrounds.
%}}}2
%}}}1

%{{{1 music about music
\section{Music about Music in the Villancico Genre}

%{{{2 survey, types
Sacred villancicos flourished especially in the second half of the seventeenth
century but continued to be a prominent element of Spanish Christian worship
through the nineteenth century in some places.
(The simpler folkloric Christmas carols denotated by \term{villancico} today
probably developed in parallel to this tradition of complex notated music and
absorbed some influences from it.)
Communities from Madrid to Manila heard and performed villancicos on the
highest feast days of the year---not just Christmas, but also Epiphany, Corpus
Christi, the Conception of Mary, and saints' days of local importance.
(They were less common in Holy Week, but there examples of \quoted{passion}
villancicos.)
Villancicos were typically presented in sets of eight or more, and were
interpolated after or in place of the Responsory chants of the Matins liturgy.
They were also sung in Mass and during Eucharistic devotional services.
Festival crowds heard villancicos in the public square in processions and
mystery plays, especially on Corpus Christi.

Villancicos were at once a genre of poetry and of music.
As religious lyric poetry, villancicos were printed in unbound leaflets or
broadsheet (\term{pliegos sueltos}) that advertised or commemorated the
performance of the musical settings in a particular place.%
    \citXXX[catalogs, studies of pliegos]
These imprints were disseminated all across the empire, partly through networks
of musicians (see \cref{ch:cererols-suspended}, for example).
A single \term{villancico poem} is in many cases just one individual variant of
a broader \term{villancico family}, a group of closely related texts and their
variants.

As a musical genre, the settings of villancico poems prioritize the clarity and
meaning of the words, and like other seventeenth-century vocal music typically
present one phrase of text at a time.
They were performed by ensembles of voices, from as small as one or two solo
voices to as large as three separate choruses of a dozen vocal parts with
multiple singers per part.
The very small-scale pieces were similar in texture and style to Italian or
German sacred concertos and English verse anthems; later in the seventeenth
century they become more like continuo songs.
The pieces with larger scoring feature lively contrasts of texture between each
choir, the full ensemble, and soloists, rather like an English full anthem.
Voices were typically accompanied by a continuo group of bass-line instruments
like dulcian (\term{bajón}) and polyphonic instruments like harp and organ.
Vocal parts could be doubled with reed instruments like dulcians and shawms
(\term{chirimías}) and brass instruments like sackbuts (\term{sacabuches}).

The words for these pieces were in Spanish and sometimes other vernacular
languages including Portuguese, Catalan, and Náhuatl, along with pieces whose
texts imitated the dialects of African slaves.
The music varied in style and technique from elements of common dances
and popular tunes up to the most sophisticated polyphonic tone-painting.
Sets or cycles of villancicos for a particular feast like Christmas included
many different subtypes of villancicos within them, offering something for
everyone.

The structure of villancicos reflects the effort to communicate on multiple
levels.
The \term{estribillo} or refrain section of a typical villancico was presented
by the full ensemble at the beginning and then repeated at the end; composers
usually set this in relatively complex polyphony similar to what they would use
for a motet.
In the center of the piece, the \term{coplas} or verses were usually set
strophically for solo singers or a reduced ensemble with accompaniment.%
\begin{Footnote}
    Sometimes preceding the estribillo was an \term{introducción}, usually for
    a smaller group; and in earlier examples the estribillo could lead to a
    \term{responsión}, an amplified version of the estribillo for full
    ensemble.
\end{Footnote}
Bernardo Illari argues that the coplas spoke more directly to common people.%
    \Autocite{Illari:Polychoral}
It would have been easier for them to make sense of the words that were sung to
the simple, repeating melodies.
The tunes may also have been familiar if they were based on oral traditions for
singing poetry to stock melodic formulas, especially in \term{romance} meter.
The \term{estribillo}, by contrast, would appeal more to the cultivated elite,
as it is often much more complex and draws on traditions of learned
counterpoint.

Our initial examples of metamusical villancicos by Gutiérrez de Padilla and
Cererols combine several of the common tropes of \quoted{music about music} in
the villancico genre, as evidenced by a global survey of extant villancico
poems and music.%
\begin{Footnote}
    The survey was based on the listings in catalogs and published studies and
    from archival sources, some previously unknown, from all over the former
    Spanish Empire (see \cref{biblio}).
\end{Footnote}
More than eight hundred villancicos were found in which themes of musical
hearing were central, a number that only hints at the original size of this
repertoire.
These metamusical villancicos may be grouped in eight main categories
(\cref{tab:survey}): in descending order of frequency these are hearing and
sound, music and singing, birdsong, dance, musical instruments, angelic
musicians, music of the heavenly spheres, and pieces that treat the
relationship of sensation and faith.
An additional category of pieces about affects is also included, since these
pieces, though not explicitly about music, do address the question of how
listeners should respond in worship. 

%{{{5 table survey topics
\insertTable{survey}
{Topics of metamusical villancicos in global survey}
%}}}5

In each of the categories in \cref{tab:survey}, we may distinguish between two
main ways of referring to music.  
Some pieces are primarily imitative, referring to real human music-making
(Boethius's \term{musica instrumentalis}).
These pieces are highly \term{intermusical}, in the way a verbal text full of
references to other texts is intertextual.
In contrast to this first category of imitative pieces, villancicos in a second
category refer to music as more of an abstract concept, such as the higher
Boethian levels of music, music as a Neoplatonic ideal, and the music of
Heaven---notions that overlap in inconsistent ways in early modern thought.
Of course, the pieces in the latter group still refer to music in the abstract
through the medium of real sounding music.  
Some of these pieces depend more on \term{intramusical} relationships---that
is, musical references internal to the individual piece itself, such as melodic
or rhythmic motives or internal contrasts of musical style without overt
references to pre-existing styles \quoted{outside the piece}.
In this section we will look at key examples in several of these categories,
moving up a Neoplatonic chain of being from the simplest imitations of birdsong
to the more conceptually challenging evocations of heavenly music.
%}}}2

%{{{2 imitative references
\subsection{Imitative References to Music: Birdsong, Instruments, Songs and
Dances}

%{{{3 birdsong
A frequent example of imitative musical reference in villancicos is when the
ornamented vocal lines are used to represent birdsong.
In a piece called \wtitle{Sagrado pajarillo} (Little sacred bird), Zaragoza
composer José de Cáseda sets the lyrics \foreign{con gorgeos} (with trills) to
twittering melismas (\cref{mux:CasedaJ-Sagrado_pajarillo}).%
\begin{Footnote} 
    This piece comes from the archive of the Conceptionist Convento de la
    Santísima Trinidad in Puebla de los Ángeles and is now preserved at CENIDIM
    in Mexico City (\sig{MEX-Mcen}{CSG.155}).
    As with many other music manuscripts in that collection, the original
    lyrics (beginning \foreign{Sagrado pajarillo}) were replaced by another
    text (beginning \foreign{Fecunda planta viva}), which was pasted and sewn
    over the original words with thin strips of paper.  
    The original may still be seen by lifting the strips.
    For another example of the bird trope by this composer, see
    \cref{ch:zaragoza}.
\end{Footnote}
Birdsong had theological importance as the paradigm of music-making in the
natural world.
The widely read Dominican writer Fray Luis de Granada, in his theological
interpretation of the natural world, says that birds reflect the harmony built
into the created world by its divine Creator because they sing purely by nature
rather than by reason, as humans do.%
    \citXXX[luis]
The only artifice to be heard in birdsong was that of God himself.
Using human voices to imitate birdsong, then, prompted listeners to consider
how the artifice of human music reflected the order of creation (this theology
is discussed in more detail below). 

%{{{5 music CasedaJ Sagrado pajarillo
\insertMusic{CasedaJ-Sagrado_pajarillo}
{Bird-like trills in Cáseda, \wtitle{Sagrado pajarillo}, excerpt from the
estribillo, Tiple I-1}
%}}}5
%}}}3

%{{{3 instruments: percussion
Next to the musical sounds of animals, the sounds of musical instruments
provided rich possibilities for musical imitation in a theological context.
Wooden sounding boards, brass pipes, and gut strings allowed human players to
take the potential of music built into the created world---such as the perfect
Pythagorean ratios of the overtone series---and actualize them in sound.

To imitate percussion instruments, for example, villancico composers paired
onomatopoetic nonsense words with distinctive rhythmic patterns.
Juan Gutiérrez de Padilla had the chorus of Puebla Cathedral represent the
sound of the castanets and tabor with contrasting onomatopoetic rhythmic
patterns on the words \foreign{al chaz, chaz de la castañuela, y el tapalatán
de el tamboríl} (\cref{mux:Padilla-Alto_zagales-chaz}).
Such pieces about instrumental music imitate the instrument itself while also
playing with a stylistic topic associated with that instrument.

%{{{5 music GdP Alto zagales chaz
\insertMusic{Padilla-Alto_zagales-chaz}
{Gutiérrez de Padilla, \wtitle{Alto zagales de todo el ejido}
(\sig{MEX-Pc}{Leg. 2/1}, Christmas 1653), estribillo: Imitation of castanets
and tabor (or tambourine?)}  % XXX check and signature
%}}}5

The same instrumental trope appears in a villancico poem performed at Toledo
Cathedral in 1645.%
    \footnote{\sig{E-Mn}{VE/88/12, no. 6}.}
Though the music, credited in the poetry imprint to Vicente García, has not
been found, the words alone conjure up a racket of percussion sound:
\begin{quotepoem}
    Porque los instrumentos sonaban así, 
        & Because the instruments sounded like this: \\
    El Atabal, tan, tan ,tan,	    & the drum, tan, tan, tan, \\
    El Almirez, tin, tin, tin, 	    & the mortar, tin, tin, tin \\
    la Esquila, dilín, dilín,	    & the chime, dilín, dilín, \\ 
    y la Campana, dalán, dalán,	    & the bell, dalán, dalán, \\
    Las Sonajas, chas, chas, chas,  & the rattle, chas, chas, chas, \\
    y el Pandero, tapalapatán.	    & and the tambourine, tapalapatán.
\end{quotepoem}
The instruments in this list are simple, rustic noisemakers from everyday
peasant life.%
\begin{Footnote}
    Note that this source from Toledo spells the rattling sound \foreign{chas}
    while Gutiérrez de Padilla's manuscripts from New Spain spell it
    \foreign{chaz}.
    This reflects the different pronunciation of the Andalusian settlers of
    central Mexico, who according to phonetic spellings in the music
    manuscripts pronounced \term{ci}, \term{zi}, and \term{si} all with an S
    sound.
    In Toledo, the first two of those phonemes would start with an English TH
    sound.
\end{Footnote}
In this villancico these instruments, which are described further in the
coplas, join together with the sounds of the mule and other animals, and the
dances of the shepherds.  
This piece, like many villancicos, depicts a scene of common folk rejoicing
after their own fashion in the humble setting of the Bethlehem stable.
The focus here is not on instrumental performance in the present day but on
helping listeners imagine the sounds of the first Christmas---but one cannot
help speculating whether peasants in Toledo might have brought such instruments
with them into the church at Christmas.

Imitating an instrument, though, did not mean that the instrument was actually
used in church; indeed in many cases the situation seems to have been the
opposite.
Despite the fanciful reconstructions that can be heard in modern recordings, no
one has yet provided documentary evidence that percussion instruments were used
in church.
They appear only rarely in images of church ensembles or iconography of angelic
consorts (see \cref{fig:BMV-Montserrat,fig:Correa-Sacristy}), and known
archival records do not record payments to makers or players of these
instruments, the way they do for shawm, dulcian, organ, and harp.%
    \citXXX[evidence needed]
%}}}3

%{{{3 clarines
\subsubsection{Becoming Clarions}

Another common example of the imitative, intermusical type would be the many
pieces that mention the \term{clarín} (clarion or bugle), in which the
singers perform patterns that are meant to sound like brass fanfares.
The typical style of clarion evocations may be seen in two examples from the
archive of the Escorial, which holds much of the surviving repertoire of the
Spanish Royal Chapel.
Most clarín pieces do not actually feature written-out clarín parts; in most
cases the instrument is imitated vocally or by other instruments, like
\term{chirimías} (shawms).
Matías Durango's \wtitle{Cajas y clarines} (Drums and bugles) evokes these
instruments with voices and shawms in martial style, as part of a broader
battle topic.%
    \footnote{\sig{E-E}{Mús. 29/15}.}
Durango's clarín topic is strikingly similar to one of the rare surviving
clarín parts from a villancico, in a fragment by the prominent Madrid composer
Sebastián Durón (\cref{mux:durango-duron-clarines}). 
Both are in the same collection of music from the Royal Chapel preserved at the
Escorial.%
\begin{Footnote}
    \sig{E-E}{Mús. 29/15} (Durango), \sig{E-E}{Mús. 32/16} (Durón).
\end{Footnote}
A villancico by José Romero from about 1690, \wtitle{Suene el clarín} (Let the
clarion resound) includes an actual notated part for \foreign{los clarines de
los autos}, that is, for the clarions played in the \term{autos sacramentales}
or public Corpus Christi dramas.% 
\begin{Footnote} 
    \sig{D-Mbs}{Mus. ms. 2914}, edited in \autocite[655--661]{CaberoPueyo:PhD}.
\end{Footnote}
The sung voices layer bugle-like gestures above them, creating a more complex
fanfare than the valveless instruments could play on their own.

%{{{5 music: clarin in voice vs actual, durango/durón
\begin{musicexample}
    \label{mux:durango-duron-clarines}
    \caption{An imitation of \term{clarín} music by voice and shawm compared
    with an actual \term{clarín} part: 
    (1) Durango, \wtitle{Cajas y clarines} (\sig{E-E}{Mús. 29/15}, Tiple I-1,
    estribillo); 
    (2) Durón, \wtitle{Dulce armonía} (\sig{E-E}{Mús. 32/16}, estribillo)}

    \includeFloatPDF{Durango-Cajas_clarines}
    \includeFloatPDF{Duron-Dulce_armonia_clarin}
\end{musicexample}
%}}}5

Perhaps there are few clarín parts because these instruments may not always
have been allowed in church, or perhaps their music was generally improvised by
the class of Spanish community musicians called \term{ministriles}, similar to 
the Lutheran \term{Stadtpfeiffer}.%    
    \citXXX[stadpfeiffer, ministriles (Illari), Praetorius CD]
In any case, the instrument was more important as a symbol than as part of the
chapel ensemble.
The \term{clarín} was used in military, royal, and apocalyptic symbolism as far
back as the allegorical \foreign{clairon} fanfares in the 1454 Feast of the
Pheasant hosted by the ancestor of the Habsburg monarchs, Philip the Fair of
Burgundy.%
\begin{Footnote}
    \Autocites[340--380]{LaMarche:Memoires}{Bloxam:JNV}{Perkins:Patronage15C};
    on the symbolism of this instrument in contemporary Spanish drama, in which
    \term{Clarín} was the name of a comic stock character, see
    \autocite{Damjanovic:Clarin}.
\end{Footnote}
In \wtitle{No temas, no recelas} by another famed Madrid composer, Cristóbal
Galán (from \circa{1691}), the voices represent \term{clarín} music in a scene
of \quoted{heavenly armies} going to battle.% 
\begin{Footnote} 
    \sig{D-Mbs}{Mus.  ms. 2892}, 
    edited in \autocite[555--565]{CaberoPueyo:PhD}.
\end{Footnote}

Imitating the clarion within a battle topic was not always just a spiritual
symbol: it was often used like real bugle fanfares were, to celebrate military
victories, or boost morale in the midst of conflicts.
The anonymous villancico \wtitle{Noble clarín de la fama} states on the cover
page that it was performed \quoted{for the profession of the sisters
\foreign{Señoras} Sor Sagismunda and Sor Jacinta Perpinyà into the Convent of
Santa Clara of Gerona, 1693}.% 
    \footnote{\sig{E-Bbc}{M/772/35}.}
The surname of these siblings (sisters by blood and now by vow) is the name of
Perpignan, capital of the Catalan region of Rosselló, which had become the
French Roussillon after the Peace of the \XXX[Pyrennees] in 1659.
A long struggle over this border territory in the War of the Great Alliance
climaxed in the year this villancico was performed, as the French general
Catinat scored a major victory against the allied powers at Marsaglia.%
    \citXXX[history]
The villancico appears to align Catalan identity with the French cause, as it
praises the \quoted{Catalan Amazons, who have the name of Perpignan}, who
\quoted{seek today good protection for their defense in Francisco}---that is,
they look for protection both to Saint Francis, the probable patron of their
order, and to France.
In enlisting for spiritual battle with Francis, the estribillo suggests, the
sisters themselves are becoming clarions of war.%
\begin{Footnote}
    Excerpt from the estribillo: 
    \foreign{Noble clarín de la fama/ 
    que de vozes te alimentas,/
    toca, toca, alarma, alarma,/
    que dos niñas hoy son aliento
    de tu voz excelsa,/
    Catalanas amazonas,/
    de Perpiñan nombre tienen,/
    pues bella guardia en Francisco,/
    buscan hoy por su defensa,/
    cuidado serafines,/
    resuenen los clarines}.
\end{Footnote}

At this moment of commitment in these young women's lives, coinciding with a
political crisis, the concept of \emph{becoming} a clarion held more
significance theologically than the mere sound of the actual instrument would
have held.
This concept is realized even more completely through musical representation in
a villancico by Juan Hidalgo (1614--1685, composer of the first Spanish operas
for the royal court), \wtitle{Venid, querubines alados}.%
\begin{Footnote}
    \sig{D-Mbs}{Mus. ms. 2895}. 
    On Hidalgo's theater music, see \citXXX[Stein].
\end{Footnote}
In this chamber villancico or \term{tono divino}, the two voices sing that just
as the birds of the dawn are \term{clarines} celebrating the Blessed Virgin, so
too will their own voices become \term{clarines}
(\cref{poem:Venid_querubines_alados-Hidalgo}).
Hidalgo interweaves the two voice parts in rising fanfare gestures that
actually allowed listeners to hear the singers transforming their voices into
\term{clarines} (\cref{mux:Hidalgo-Venid_querubines}).

%{{{5 poem and music Hidalgo Venid querubines
\insertPoem{Venid_querubines_alados-Hidalgo}
{\wtitle{Venid querubines alados}, poem set by Hidalgo (\sig{D-Mbs}{Mus. ms.
2895}), copla 5}

\insertMusic{Hidalgo-Venid_querubines}
{Hidalgo, \wtitle{Venid querubines alados}, duo response at end of each copla}
%}}}5

These villancicos use the \term{clarín} as a metonym for music-making
generally, subsuming both theological and political aspects of the instrument
and its characteristic style as central to the notion of music itself.
Theologically, they summon a range of scriptural references to brass
instruments---the trumpets of King David's priestly musicians
(\scripture{IChr}{0:0}), the trumpets of the apocalypse
(\scripture{Rev}{0:0}; \scripture{IThess}{0:0})---where clarion-like
instruments were used in earthly and heavenly worship and served as signs of
God's divine authority and calls to attention at moments of God's judgment,
signals of divinely ordained seasons and times.%
    \citXXX[biblical trumpets]
At the same time, we must not overlook the obvious political significance of
clarions in the militaristic society of early modern Spain---meanings that were
also understood in Biblical terms.
Just as the biblical Jericho had fallen by the divine hand at the sound of the
trumpets of Gideon's army (\scripture{Josh}{0:0}), clarion fanfares heralded
the arrival of the Spanish monarch or viceroy, communicated to troops on the
battlefield, and generally proclaimed to the ears what flags and triumphal arches
conveyed to the eyes---the dominion and divinely sanctioned authority of the
Spanish king over his global realms.\citXXX[bible verses]

The clarion, put simply, was a sign of power.
Clarion-themed villancicos, then, depended on the instrument's signification of
power to proclaim and reinforce the sovereignty and authority of the Spanish
church and state.
Under the Spanish \term{padronado} (the Spanish monarch's self-appointed
guardianship by direct rule of the Catholic Church in his realm), church and
state power were aspects of the same governing authority, and together they
administered rewards and punishment in both temporal and eternal domains.
Spanish subjects were taught a theological concept of society in which the
world was stratified in a static hierarchy of types and stations of people, in
a way understood to be harmonious and divinely ordained.

The continued reiteration of power through music, and the conflation of
political and theological symbolism, could signify terror and oppression for
many people but also serve as a consoling reminder to others of the order and
stability of the world.
The right man was in charge, the clarion call announced: God was working
through the rulers of the world to govern his creation; common people were
protected and defended against the forces of evil.
That early modern Spaniards understood evil in the human forms of the Moor, the
Jew, the heretic Lutheran;
that they believed the divinely sanctioned order of society consigned
\soCalled{Indians} and \soCalled{Blacks} for roles of servitude; 
that they believed only males could hold authority and that
an inequitable distribution of wealth was part of the natural hierarchy---the
clarion call signified all this, too.
We can affirm that the \quoted{music of state} in the Spanish Empire served as
a \quoted{instrument of dominion}, even in some ways fulfilling functions that
may be seen as prototypes for later totalitarian propaganda, while also
acknowledging that many Spaniards and their colonial subjects actually believed
in the theological foundations of their political order and even actively
contributed to reinforcing the hierarchical power structure, including through
patronizing, performing, and listening to music.%
    \citXXX[Rodriguez, Sage, Rietbergen, Menache, etc]
%}}}4
%}}}3

%{{{3 dance, ethnic vcs
\subsubsection{Dance and Difference}

Dance topics in villancicos provided another way for the genre to point beyond
itself to other kinds of music in society, and like clarion topics these
references both reflected and reinforced Spanish attitudes toward social
structure.
Many dances are explicitly named and often the text proclaims the villancico
itself to \emph{be} a specific kind of dance, including \term{zarabanda},
\term{jácara}, \term{guarache}, \term{danza de espadas}, and
\term{papalotillo}.%
    \citXXX[sources]
Only a few of these dance references have been corroborated with other notated
sources of dance music, and while those other sources, byproducts of a chiefly
improvised tradition, are often sketchy and difficult to reconstruct, the
villancico arrangements of dances provide a complete musical texture.%
\begin{Footnote}
    Compare, for example, the elaborate instrumental dances recorded by Andrew
    Lawrence-King and the Harp Consort, \headlesscite{Lawrence-King:DancesCD}
    with the rudimentary notation of a few bars of chords and minimal strumming
    patterns in the source, \shortcite{Ruiz:Luz}.
\end{Footnote}
In many cases the villancico settings may be our only remaining traces of these
dances.%
\begin{Footnote}
    More research is needed on the question of whether performers of
    villancicos actually danced or used other elements of staging in church.
    There certainly was ritual dance on Corpus Christi in Seville and Valencia
    Cathedrals, performed by the boy choristers known as \term{seises}.%
        \Autocite{Comes:Danzas}
\end{Footnote}
	
Gutiérrez de Padilla's sacred \term{jácaras} for Puebla Cathedral in 1651--53
and 1655 all play with variants of the same basic tune, harmonic pattern, and
rhythmic groove, and are closely similar to the music Álvaro Torrente has
reconstructed for the secular \term{jácara}.%
    \citXXX[padilla, torrente]
But rather than presenting the ballad of an outlaw using underworld slang,
Gutiérrez de Padilla's sacred \term{jácaras} celebrate the child Jesus as the
boldest and most ingenious hero of them all.
In terms of musical theology, these pieces perform the meeting of disparate and
seemingly opposed elements, which was the central theme of the feast of
Christmas.
In one piece, the baby Jesus is a gunslinger \quoted{from way up in Texas}.
Another piece self-consciously acknowledges the collision of worlds at the
heart of this particular type of \quoted{metaphorical twist}: in the
\term{jácara}, the chorus says, \quoted{we present as a dish of the court what
really is feed from the village}.%
    \citXXX[sources]
In these terms, the use of the song/dance tradition mixes not sacred and
secular but elite and common, refined and crude, the appropriate and the
\quoted{bizarro} (a common term in \term{jácaras}).

Later \term{jácaras}, though, lose many of the explicit connections to their
worldly origins and become another conventionalized subgenre of villancicos.
Mateo de Villavieja's \term{Jácara en anagramas} continues the concept of
ingenuity from earlier \term{jácaras} as it plays with permutations of a set of
textual and musical elements (an early example of algorithmic composition), but
it retains none of the musical or poetic elements of earlier \term{jácaras}.
Sacred jácaras became yet another pious entertainment for the educated classes,
perhaps with a bit of added thrill by their association with ribald origins,
the course feed of the village trough having been by this time almost
completely refined away into purely a dish of the court.

Metamusical references to traditional music-making of lower-class people
extended also to the depiction of ethnic difference.
There are villancicos that depict non-Castilian groups like Native Americans,
African people, Catalans, Frenchmen, even Irishmen, through caricatured
deformations of language and music.
What have come to be called \quoted{ethnic villancicos} were labeled in their
time as \term{villancicos de naciones} or by the name of the particular ethnic
type for that piece, like \term{gallego} (Galician), \term{gitano}
(\quoted{Gypsy}), \term{indio} (\quoted{Indian}), or \term{negro},
\term{guineo}, and similar terms for Africans.
Most of these pieces, and especially the \term{villancicos de negro}, refer
specifically to the characteristic music and dancing of these groups, often
naming their instruments and describing their motions.
The texts use some smatterings of foreign words but mostly ask the performers
to put on an accent in Spanish: in these caricatures the Gypsy ends all her
words with a Z (\foreign{Puez los trez zon Magoz,/ hombrez de ezfera}).
\begin{Footnote}
    \term{Vamos al portal gitanilla}, Imprint from Epiphany 1666, Zaragoza,
    Iglesia de El Pilar (\sig{E-Mn}{VE/1303/1}), later attributed to Vicente
    Sánchez, \headlesscite[203--204]{Sanchez:LiraPoetica}.
\end{Footnote}
The Black says when he should say R and J, drops ending S sounds, and
mismatches genders and cases (\foreign{Mi siñol Manuele, \Dots{} Sesu, \Dots{}
pluque son linda cosa}).%
\begin{Footnote}
    \term{Al establo más dichoso}, Music manuscripts by Juan Gutiérrez de
    Padilla of \term{ensaladilla} for Christmas 1652, Puebla Cathedral
    (\sig{MEX-Pc}{Leg. 1/3\XXX}), \XXX[WLSCM32].
\end{Footnote}
Villancicos about African characters also frequently feature nonsense
syllables, whether lullaby phonemes like \foreign{ro ro ro ro} and \foreign{le
le le le}, or apparent gibberish like \foreign{tumbucutú, cutú, cutú} and
\foreign{gulumbé, gulumbá} that tells us what African languages like Kikongo
sounded like to a Spanish ear.%
    \citXXX[al establo, other]
This type of piece represents Africans as always happily engaged in drumming
and dancing.%
    \citXXX[baker etc]

These pieces were created by Spaniards primarily for other Spaniards;
\quoted{black villancicos} are not really about depicting African identity but
are rather ways of constructing a Spanish one by reference to the Other.
Immediately after purging Iberia of both Moors and Jews, the Castilians had
been overwhelmed with encounters with new ethnic groups, languages, and
religions around the world; these pieces offered the potential to create an
imagined world in which all these groups were situated in their proper place
within a well-controlled social hierarchy.
These pieces may offer glimpses of the language and music of non-Spanish
groups, but only through a glass heavily darkened by racial prejudice and
deliberate caricature for the sake of humor and mockery.

With regard to the nature of musical references in \soCalled{ethnic}
villancicos, then, these pieces encompass a mixture of literal imitation (as of
percussion, and of the \soCalled{musical} sound of foreign languages) and
broader stylistic references (as, perhaps, to African musical styles, though no
one has yet demonstrated concrete evidence for a connection).
They also include more abstract references to music through the use of nonsense
words that, somewhat like solmization syllables (see below), symbolize and
enact music-making.
Like \term{jácaras}, ethnic villancicos grow increasingly conventionalized over
the years and more distant from the low-caste sources they grew from, so that
the \term{negro} character in one year's villancicos was much more similar to
the \term{negro} of the previous year's set that he probably was to any real
person of African descent.
And like \term{clarín} pieces, ethnic villancicos both reflected and reinforced
imperial Spain's power structure by projecting a theological vision of that
structure as divinely ordained and immobile.

Their discourse on racial difference must be understood within a Neoplatonic
theological concept of music and society, in which the lowliest elements in the
created world could lead a person to the knowledge of the highest.
Juan Gutiérrez de Padilla, who included a \quoted{black villancico} in most of
his Christmas cycles for Puebla, depicted the paradox of Neoplatonic thought
when in 1652 he had his black characters, described as Angolans in the piece
like Gutiérrez de Padilla's own slave Juan, say \quoted{Listen, for we are
singing like the angels}.
As the Angolans go on to sing a vernacular \term{Gloria in excelsis} in their
dancing triple meter, full of syncopations notated by coloring in the mensural
noteheads, above them suddenly enter the two boy soprano parts of the second
chorus, which have been silent until now, singing the same \term{Gloria} with
them---but in the white notes of duple meter, and quoting a plainchant
intonation (\cref{mux:Padilla-Al_establo-Gloria}).
The Angolans and the angels are brought together for a miraculous moment
through contrasting types of rhythmic movement in which the hidden harmony
between earthly and heavenly music is revealed.
The Angolans are in some ways depicted sympathetically, as instead
of the gold, frankincense, myrrh of the magi (one of whom was portrayed on
Puebla's high altar as a black African), bring the Christ-child the homelier
and more practical gifts: a potato, a toy, and diapers.
But nothing about the piece really exalts the African characters in any way
that would affect the lives of real Africans like Gutiérrez de Padilla's slave.

%{{{5 music GdP Al establo Gloria
\insertMusic{Padilla-Al_establo-Gloria}
{Gutiérrez de Padilla, \wtitle{Al establo más dichoso (Ensaladilla)} 
(\sig{MEX:Pc}{Leg. 2/1}, Puebla Cathedral, Christmas 1652), \term{Negrilla}
section: polymetrical \term{Gloria} of Angolans and angels}
%}}}5

It is possible, though more evidence is needed, that the vogue for black
villancicos at Christmas and Epiphany was linked to the practice across the
Spanish and Portuguese Empires of \quoted{Black Kings} festivals, in which
confraternities of enslaved and free people of African descent elected a mock
royal court and paraded them around their city with music and dancing, usually
with military elements with origins in the Christian Kingdom of Kongo before
the start of slavery.%
    \citXXX[sources]
This possible connection does not mean that these villancicos express any real
African voice or viewpoint; rather, they tell us about the insecurities, fears,
and prejudices of Spaniards and may help us understand how they justified their
place in an unjust society by appeal to theology and aesthetic beauty.
Gutiérrez de Padilla's polymetrical Gloria fits perfectly with the image of
angels singing and dancing on the round painted high the new Puebla
Cathedral's high altar, hovering over the images of shepherds and magi greeting
the newborn Christ at the altar's base (see \cref{ch:padilla-voces}), and we
can imagine the theological aesthetics of this were some comfort to this
chapelmaster-priest and his peers; but they were no help to enslaved men and
women and, when such pieces are revived uncritically today, they continue to do
their historic work of reinforcing racial prejudice and appeasing the
consciences of elite (and typically white) listeners.

With regard to hearing and faith, ethnic villancicos and black villancicos in
particular enabled Spaniards to envision themselves as the rightful masters of
a society in which other groups were naturally subordinate; in other words what
they heard helped them believe in the rightness of the social order as governed
by the church. %XXX
Though their representations are purposefully distorted, they do suggest that
the Spanish elite accepted the coexistence of multiple languages and types of
music within society, and enjoyed sampling these exotic sounds through the safe
filter of their own caricatures.
%}}}3

%{{{3 VCs about VCs
\subsubsection{Villancicos about Villancicos}

%{{{4 overview
The conventions of the villancico genre itself become the subject of a special
type of self-referential villancico \emph{about} villancicos.
In one sense, the many pieces beginning \quoted{Listen} or \quoted{Pay
attention}, might be considered self-referential, since in these pieces the
singers usually announce something about the piece, as in the setting of
\wtitle{Oigan, oigan la jacarilla} by José de Cáseda, or the poem performed in
Montilla in 1689, \wtitle{Oíganme cantar una tonadilla}---\quoted{hear me sing
a tonadilla}.% 
    \begin{Footnote}
    Respectively, \sig{MEX-Mcen}{CSG.151}, \autocite[116 (no signature
    listed)]{BNE:VCs17C}.
    See \XXX[LeGuin:Tonadilla].
    \end{Footnote}
This posture rhetorical posture owes something to the genre's close
relationship with the psalms in Matins, which are filled with such
self-referential statements (\quoted{Sing to the Lord a new song},
\scripture{Ps}{97:0}; \quoted{Come, let us worship and bow down},
\scripture{Ps}{95:0}).
But it also draws on the comic and satirical elements of the Spanish minor
theater, the low-register plays (\term{entremeses}) performed between acts of
the more highbrow \term{comedias} by, for example, Lope de Vega and Calderón.%
    \citXXX[entremeses]
Similar to the Italian \term{intermezzi} skits of the eighteenth
century that were the cradle of comic opera, Spanish \term{entremeses} were
built out of formulaic scenarios and stock characters---many of the same ones
like Gil, Pascual, Bras, and Bartolo who appear in villancicos---and parodied
the conventions of the \term{comedia}.

The patrons and creators of villancicos developed well-reinforced expectations
in their audience not only for different types of villancicos, but also
possibly for the dramatic shape of the whole villancico cycle.
The surviving musical settings of complete cycles, such as those in Puebla and
Segovia, demonstrate a range of expected musical conventions for each of these
subgenres.
Miguel de Irízar's requests to his chapelmaster peers for more
\quoted{villancicos de chanza}---comic villancicos---suggests the need to fill
out each villancico cycle with certain types of pieces, mixing serious and
comic subgenres.%
    \Autocite[78]{Olarte:Irizar} 
When a villancico represents the performance of a villancico and commentary on
it, we are offered a glimpse of how people listened to villancicos.
%}}}4

%{{{4 ex: Anton Llorente
The anonymous villancico \wtitle{Antón Llorente y Bartolo}
(\cref{mux:Anton_Llorente}) presents two characters from a well-known
\term{entremés} with close links to Cervantes' \wtitle{Don Quijote}, who want
listeners to hear out their complaint about villancicos.
The villancico poem is found in a 1639 Christmas imprint from Toledo Cathedral
and an anonymous musical setting survives from the Convento de la Santísima
Trinidad in Puebla.%
\begin{Footnote}
    \sig{E-Mn}{VE/88/6}, \sig{MEX-Mcen}{CSG.014}.
    The title has been erroneously \quoted{corrected} in the Sánchez Garza
    catalog to \wtitle{Anton, Lorente y Bartolo} despite the clear double L in
    the manuscript (which was never used when a hard L sound was
    intended).\XXX[check]
\end{Footnote}
The more well-known stock characters Gil and Bras, they say, have held the
stage for too long:
\begin{quotepoem}
    Antón Llorente y Bartolo	& Antón Llorente and Bartolo \\
    trazaron un memorial	& drew up a complaint \\
    de que con los villancicos	& that with all the villancicos \\
    se han alzado Gil y Bras.	& Gil and Bras have gotten the spotlight.
\end{quotepoem}
Anton Llorente and Bartolo insist that they could make a good enough villancico
of their own if given the chance:
\begin{quotepoem}
    Si ha de sonar el pandero,	& If the tambourine is going to be played, \\
    solo Gil le ha de tocar,	& it is only Gil who ever plays it, \\
    y si ha de haber castañetas,& it if there have to be castanets, \\
    ha de repicarlas Bras.	& Bras is the one to rattle them. \\
    También acá somos gentes	& But here we are, we too are good fellows, \\
    y alcanzar podemos ya	& and we can even manage \\
    de un villancico un bocado	& a nibble of a villancico \\
    y un pellizco de un cantar.	& and a pinch of a song.
\end{quotepoem}
In the succeeding \term{responsión} section, the full eight-voice chorus joins
in endorsing the new characters and denouncing the old:
\begin{quotepoem}
    No quiero que me Brasen y que me Gilen 
    & I don't want them to Bras me or Gil me \\

    sino que me Llorenten y me Toribien. 
    & but only to Llorente me and Toribio me.
\end{quotepoem}

The anonymous musical setting for this embodies all the stereotypes of the
villancico genre, first encountered here in Gutiérrez de Padilla's \wtitle{En
la gloria de un portalillo}.% 
\begin{Footnote}
    One possible composer is the Seville Cathedral chapelmaster Fray Francisco
    de Santiago, whose setting of this text was cataloged as part of the
    now-lost library of King John (João) IV of Portugal (see
    \cref{ch:padilla-voces}), 
    \autocite[caixão 26, \range{no}{675}]{JohnIV:Catalog}.
\end{Footnote}
The piece is in highly accented triple meter (\meterCZ) with continual use of
sesquialtera, and it opens with a declamatory section for full chorus, followed
by a vocal solo that is then echoed in polychoral dialogue by the full
ensemble.
The text is both dramatized and symbolized by leaping gestures that leap from
voice to voice in points of imitation on \foreign{salten y brinquen} (jumping
and leaping).
These features may just constitute typical villancico style, or they may be
taken to \emph{represent} typical villancico style.
The highly conventional music casts the anticonventional text into relief while
also dramatizing the scene, since the piece is meant to portray Anton Llorente
and Bartolo performing a villancico.
This is a villancico, then, in the style of villancicos.

% TODO	See also in lafragua connection; uber-conventional villancico next to
% anticonventional one

% TODO is there a Lisbon Anton Llorente?

%{{{5 music Anton Llorente
\insertMusic{Anton_Llorente}
{Anonymous, \wtitle{Anton Llorente y Bartolo} (\sig{MEX-Mcen}{CSG.014}), first
stanza of introducción and beginning of responsión (Accompaniment omitted)}
%}}}5

As though the 1639 Toledo text were not self-referential enough, the creative
team at the cathedral followed up the next year with another villancico that
specifically referred back to \wtitle{Anton Llorente y Bartolo}.%
\begin{Footnote}
    \ptitle{Quejosos de la sentencia que dio el alcalde Pasqual}, in imprint
    from Christmas 1640 at Toledo Cathedral, \sig{E-Mn}{VE/88/7,
    \range{no}{2}}.
\end{Footnote}
The narrator says that the \foreign{Brases} and \foreign{Giles} were so
\quoted{frustrated by the sentence that Mayor Pasqual decreed against them last
Christmas}, that \quoted{they appealed to another one who was more learned}
(the \quoted{Mayor of Bethlehem} was another stock character in comic
villancicos).
Each one states his case for why he is needed at the Nativity, and Bras's
conclusion wryly sends up the conventionality of villancico poetry:
\begin{quotepoem}
Cuanto ha qué Belén lo es,	& As long as Bethlehem has been what it is, \\
y ha sido el portal portal,	& and the stable has been a stable, \\
a peligros de poetas		& where poets have been in danger, \\
ha sido socorro Bras.		& Bras is always there for aid.
\end{quotepoem}
The new mayor, in the name of keeping traditions, undoes the sentence of the
previous year, and the chorus rejoices, because without Bras and Gil it would
not be Christmas:
\begin{quotepoem}
Que me Brasen, y Gilen	& I wish them \\
quiero zagales,		& to Bras me and Gil me, lads, \\
porque no soy amigo	& because I am no friend \\
de novedades.		& of novelties.
\end{quotepoem}
The chorus, speaking here for the mayor's subjects in the community, affirms
the decision to keep their familiar Christmas characters:
\begin{quotepoem}
Porque en saltando a esta fiesta & For if you take from this feast \\
el pesebre, y el portal,  	 & the manger, the stable, \\
las pajas, Brases, y Giles, 	 & the straw, Brases, and Giles, \\
no es fiesta de Navidad.	 & it is no festival of Christmas.
\end{quotepoem}
Here we have a scene of people clamoring for villancicos with all their corny
conventions as a central part of making Christmas feel like Christmas.

As the mayor says, one reason villancicos were so conventional may be because
the feast they were most closely associated with was (and is) one where customs
are carefully preserved.
Novelty at Christmas is expected but only within certain traditional
boundaries.
Part of cultivating those customs meant naming them explicitly in song, as we
have already seen in Cererols's \wtitle{Fuera que va de invención} and several
other pieces, like a North American Christmas tree ornament in the shape of a
Christmas tree.
Comic villancicos like these should not be written off as less theologically
motivated than the more cultivated pieces.
Though the \wtitle{Anton Llorente} pieces present no learned theological
doctrines, they still serve a religious function in prompting hearers to
laughter and enjoyment, and that function contributed to the effect of a set of
villancicos within the liturgy.
The comic pieces may even have been the most likely to provoke direct responses
of laughter and surprise in many listeners, and therefore could be the most
effective in actually moving those assembled toward sympathetic vibration and
harmony together. 
They also served the practical goal of attracting and pleasing parishioners and
making them feel at home in the church---a purpose that might even be
considered more truly theological in the sense of fulfilling the church's
religious purpose as a faith community rather than projecting theological
concepts.%
\begin{Footnote}
    Contemporary Catholic theologian David Fagerberg argues that the first
    and foundational meaning of theology as an activity---what he calls
    \term{theologia prima}---is what the simple worshipper does during the
    liturgy.
\end{Footnote}
%}}}4
%}}}3
%}}}2

%{{{2 abstract references
\subsection{Abstract References to Music as Concept or Symbol}

%{{{3 overview
In the second category of metamusical villancicos are pieces that refer to
music more as an abstract concept, rather than to a specific, identifiable
reference to another kind of music.
When Pedro Ruimonte in \wtitle{Gil, pues a cantar} (one of the few villancico
settings to be published in print) sets the word \foreign{cantar} (sing) to a
long melisma, or when Gaspar Fernández in \wtitle{Sobre vuestro canto llano}
illustrates the phrase \term{canto llano} (plainchant) with imitative
counterpoint around a cantus-firmus-like Tenor part, these composers are using
the characteristic emblems of vocal music to refer to the concept of singing in
general.%
    \citXXX[ruimonte, fernandez]
%}}}3

%{{{3 solfa
\subsubsection{Solmization Puns and the Theology of Voice Alone}

One of the most common ways of explicitly singing about singing was to use
solmization syllables---\term{ut, re, mi, fa, sol, la}---in the poetry.
References to Christ as \term{sol} (sun) are ubiquitous, and as shown in the
opening example by Gutiérrez de Padilla, composers missed no opportunity to put
this word on a pitch that could be solmized with that syllable (G, C, or D in
the three Guidonian hexachords).
Solmization tropes brought the rudiments of musical artifice into the
foreground, forcing educated listeners to take note of the constructed nature
of what they were hearing.
No pun was too obvious: in composer Miguel de Aguilar's \term{oposición}---his
audition piece---for a position at Zaragoza, \wtitle{Mi sol nace y tiembla}, it
is not hard to guess the opening pitches he chose.%
\begin{Footnote}
    \sig{E-Zac}{B-11/233}, \ptitle{Villancico de Oposición en Zaragoza}, edited
    in \autocite[34--64]{Ezquerro:MME55}.
\end{Footnote}
Solmization syllables were sometimes used for their own sake, without a
symbolic meaning, somewhat like the \quoted{fa la la} refrains in contemporary
English madrigals.
Here sign and signified become one: the voice bears no message except the
musical voice itself.

Passages of self-conscious solmization are not alluding to a particular kind of
song; rather, their song is pointing to the abstract category of
singing in general.
In Aguilar's \wtitle{Mi sol nace}, the words have dual function: on one side
they communicate linguistic meaning (\quoted{my sun}, which is itself
metaphorical), but on the other side these musical syllables go beyond
language, to both symbolize and embody music-making. 
Aguilar made this obvious gesture at the beginning of a piece intended to
demonstrate his own skill at composition, in keeping with the tradition we will
trace in \cref{part:unhearable-music} of Spanish composers using metamusical
villancicos to establish their compositional pedigree.

At the same time, the syllables themselves could also take on deep symbolic
meanings, as in the Christological \quoted{sign of A \term{(la, mi, re)}} in
Gutiérrez de Padilla's 1657 \wtitle{Voces, las de la capilla} (the subject of
\cref{ch:padilla-voces}).
The syllables as an ordered series---that is, as a scale---served a symbolic
purpose with roots in medieval tropes like Mary as the \term{scala
peccatoris}.%
    \citXXX[Gilbert?]
Andrés Lorente began his 1672 music treatise with a dedication to the Virgin
Mary and a warning to the reader that moral virtue was the key to being a true
musician; he uses the six syllables of the hexachord as a ladder ascending to
heaven, where each syllable matches the start of a verse of Scripture
pertaining to a particular virtue, and as a second ladder descending to
hell\XXX[check], with each syllable denoting a particular vice.

The system of three hexachords offered additional symbolic potential.
Spanish music pedagogy preserved the widespread medieval system in which were
three groups of the six syllables from \term{ut} to \term{la}: the
\quoted{natural} hexachord (\term{cantus naturalis}\XXX[check]) starting on C;
the \quoted{hard} hexachord (\term{cantus durus}) starting on G and featuring
the \quoted{hard} or square B (ancestor of the natural sign \na{}); and the
\quoted{soft} hexachord (\term{cantus mollis}) starting on F and featuring the
\quoted{soft} or round B (ancestor of the flat sign \fl{}).
A singer moved or mutated between these three sets at certain prescribed
positions (for example, shifting from D as \term{re} in the natural hexachord
to \term{la} in the soft hexachord. % XXX cut all this?
The hexachords could thus be understood as transpositions of each other.
The practice of \term{musica ficta} allowed further transformations in order
to sing notes \quoted{outside the hand} (not included on the mnemonic
device of the Guidonian hand): in certain situations (e.g., going one note
above \term{la} and back again), one sang \term{fa} or \term{mi} instead of the
normal syllable to create a semitone where there would not otherwise be one.
This much music theory was part of the basic training of every cathedral
chorister and university-educated Spanish subject.%
    \citXXX[sources] % XXX distinguish cantus mollis as "key signature"

In 1678, Segovia Cathedral chapelmaster Miguel de Irízar began the festivities
of Christmas with the \term{calenda} piece, \wtitle{Qué música celestial}, in
which he used the hexachordal system to depict heavenly music coming down to
earth (\cref{mux:Irizar-Que_musica_celestial}).%
    \footnote{\sig{E-SE}{18/36}.}
The piece dramatizes the moment when the shepherds of Bethlehem first heard the
music of the angelic choir (\scripture{Lk}{2:0}) by having three speakers ask,
in turn,
\quoted{What celestial music is this which alters the air?}
\quoted{What sovereign harmony is this which elevates hearing?}
\quoted{What light is this that transforms the dense night into day?}%
    \citXXX[wlscm32+]
Of course, as the first villancico of Christmas heard in Segovia Cathedral that
year, \quoted{this music} also refers to the music currently resounding under
the massive vaults of the last Gothic cathedral in Europe. % XXX
Irízar gives the first voice to a boy soprano, who sings down all the steps of
the soft hexachord from \pitch{D}{5} (\term{la}) to \pitch{F}{4} (\term{ut}).
% XXX check voice parts
The figure is an epitome of music itself, a textbook example of solmization
that begins at the very top of the Guidonian gamut (the second highest note on
the hand).
The second voice, then, imitates the first phrase exactly, but transposed down
a fourth into the natural hexachord (from \pitch{A}{4} to \pitch{C}{4}).

%{{{5 music Irizar Que musica celestial
\insertMusic{Irizar-Que_musica_celestial}
{Irízar, \wtitle{Qué música celestial} (\sig{E-SE}{18/36}, Segovia Cathedral,
Christmas 1678), Opening}
%}}}5

The shift from the \quoted{altered} soft hexachord, with B flat, to the plain
natural hexachord, symbolizes the movement of music from heaven to earth.
% XXX other examples of cantus mollis for heaven
Between the two singers, the two phrases outline the plagal ambitus of the
second mode (from A to A, with a final on D), thus presenting hearers with a
paradigm of perfect music, according to the most ancient of rules known to a
late-seventeenth-century Spanish chapelmaster.
Meanwhile, the bass line for the continuo accompaniment adds a further
heaven--earth contrast, as it moves in canon with the singers but with a
rhythmic displacement so that the bass and melody voices form a chain of 7--6
suspensions.
The way the bass voice moves at a delay from the solo voice suggests the way
earthly music imitates or echoes heavenly music.
On the other hand, the contrapuntal pattern is a textbook example of
fourth-species counterpoint, so it could also be a way of representing heavenly
music itself.  
This kind of heavenly music defies human expectations but is at the same time
governed by its own laws.
A listener untrained in counterpoint might only have perceived a mysterious,
haunting affect, and in any case the passage does evoke a \foreign{soberana
armonía} that \quoted{elevates the sense of hearing} or \quoted{lifts up the
ear}.

For a virtuosic demonstration of solfa composition we turn again to Juan
Gutiérrez de Padilla.
In a separate part of the archive from Gutiérrez de Padilla's other villancicos
or his Latin-texted polyphony is a small handwritten notebook containing an
anthology of exemplary works gathered by an unknown musician, apparently a
student to judge from the immature handwriting.
After selections from Palestrina among others is a set of \quoted{Villancicos
of various authors}, in which there is copied just the tenor part of a setting
by Gutiérrez de Padilla, \wtitle{Miraba el sol el águila bella}.%
    \citXXX[signature, check info, esp. attribution]
The copyist had good reason to preserve this part.
Here is a villancico with a text (for the estribillo and responsión) comprised
almost entirely of solmization syllables: \foreign{Cuando mire al sol, el sol
la mire} \XXX[etc, check].
Gutiérrez de Padilla sets every syllable to the corresponding pitches so that
for much of the piece, singing the lyrics is almost identical to singing the
solmization (\cref{mux:Padilla-Miraba_el_sol-estribillo}).%
\begin{Footnote}
    We will see a similar treatment of solmization in \cref{ch:padilla-voces}.
\end{Footnote}

%{{{5 music GdP Miraba el sol estribillo
\insertMusic{Padilla-Miraba_el_sol-estribillo}
{Gutiérrez de Padilla, \wtitle{Miraba el sol}, extant Tenor part from
manuscript anthology (\sig{MEX-Pc}{Leg. 34}), estribillo: Melody matching
solmization syllables in text (Abbreviations for hexachords: \term{NAT},
\term{naturalis} on C; \term{MOL}, \term{mollis} on F; \term{DUR}, \term{durus}
on G; \term{FIC}, \term{ficta} alteration}
%}}}5 %XXX shorter caption?

% XXX avoid rep
But the piece is not nonsense---in fact, the poet (perhaps Gutiérrez de Padilla
himself) has managed to craft a semantically and theologically coherent text
based on an entirely separate conceit relevant to the feast of the Conception
of Mary, that of the Virgin Mary as an eagle.
The eagle, Spaniards believed, had the power to look directly at the sun
without harming its eyes, and thus the eagle was a fitting symbol for Mary as
Immaculate.%
    \citXXX[sources eagle]
That Mary was conceived without original sin was enforced as official dogma in
Spain long before it received papal approval in the first Vatican Council in
\XXX[18XX], and Puebla Cathedral was dedicated to Mary as Immaculate.
Gutiérrez de Padilla takes advantage of the hexachordal system, which means
that there are usually three possible notes that could be solmized with a
particular syllable, to add an additional symbolic layer embodying the eagle
conceit through musical technique.
To represent the eagle turning to the sun (and therefore Mary seeing the face
of God\XXX[?]), Gutiérrez de Padilla has the Tenor shift from the soft
hexachord, through a ficta alteration, into the natural hexachord
(\cref{dia:Padilla-Miraba_el_sol-hexachords}).
He thus quite literally moves \foreign{al sol}---both because he moves to a
note on \term{sol} and because he shifts to the hexachord that starts on the
\term{sol} of the previous hexachord.
Where Irízar shifted from the soft hexachord down to the natural for moving from
heaven to earth, Gutiérrez de Padilla makes the same shift upwards to represent
the eagle/Mary looking up to the heavens.
It is hard to imagine a more complete epitome of the concept of singing about
singing.

%{{{5 dia GdPadilla Miraba el sol hexachords
\insertDiagram{Padilla-Miraba_el_sol-hexachords}
{Gutiérrez de Padilla, \wtitle{Miraba el sol}, estribillo: Hexachordal shift
symbolizing Mary/Eagle in \term{mollis} turning to
Christ/Sun (\term{sol}) in \term{naturalis}}
%}}}5

Even when solmization passages seem to lack lexical or symbolic meaning, they
bear a profound theological meaning as an embodiment of the voice itself,
within the Neoplatonic system (explained more fully below).
In the prevailing Catholic understanding of the world, every created thing,
simply by being itself, reflected the nature of God, its Creator.
The human body was the microcosm, reflecting in turn both the whole Creation
and the Creator who took on a human body in Christ.
The voice emanated from the body and expressed the essence of the one speaking
or singing to another who heard it.
So the voice itself had quasi-sacramental meaning as an expression of Man the
microcosm and a reflection of the Creator; and this meaning was independent of
linguistic communication or even of music's own non-linguistic structures,
which were understood by analogy to verbal rhetoric.
Far from \quoted{signifying nothing}, as in the \quoted{aesthetics of pure
voice} that Mauro Calcagno identifies in the contemporary Venetian theater
productions of the Accademia degli Incogniti (far less modern philosophical
notions of the \quoted{voice itself} as separate from meaning), wordless
passages in villancicos come nearer to signifying \emph{everything}.%
    \citXXX[Calcagno; {Barthes:GrainOfVoice}{Dolar:Voice}{Cavarero:Voice}] 
    %+ TODO Feldman ed volume
%}}}3

%{{{3 music itself as a conceit
\subsubsection{Music Itself as a Conceit}

Solmization villancicos should be understood as a subtype of a category of
villancicos in which music itself is the central conceit---not a specific type
of music (human, animal, or angelic), but music in the abstract.
Such pieces often play on technical musical terms using the technique of
\term{conceptismo} to create a double discourse about both music and theology.
The most renowned of villancico poets today, Sor Juana Inés de la Cruz
(1651--1695), used the conceit of Mary as a heavenly chapelmaster to create
such a piece for the feast of the Assumption in Mexico City, 1676, though no
musical setting survives.% 
    \Autocite[\range{no}{220}, \range{p}{7}]{SorJuana:VC} 
The estribillo exhorts congregants to listen for Mary's voice
(\cref{poem:Silencio_Maria-Sor_Juana}).
In the second copla Sor Juana uses the hexachord to trace the arc of Mary's
exaltation from Annunciation to Assumption. 
Mary begins at \term{ut} with her response to the angel that she will bear
Christ (\term{Ecce ancilla}, \bibleverse{Lk}{1:46-55}), and ends on the
\term{la} of the verse that recurs throughout the liturgies for the Assumption,
\wtitle{Exaltata est sancta Dei Genitrix/ Super choros Angelorum ad caelestia
regna} (The holy bearer of God has been exalted above the choirs of angels to
the heavenly realm).
Copla 4 uses the dual identity of B as either natural (\term{mi} in the hard
hexachord) or flat (\term{fa} in the soft hexachord or in ficta) as a symbol of
Christ's dual nature as both human and divine, united in Mary's womb.

%{{{5 poem sor juana silencio
\insertPoem{Silencio_Maria-Sor_Juana}
{Sor Juana Inés de la Cruz, \wtitle{Silencio, atención, que canta Mariá},
excerpts}
%}}}5

Sor Juana demonstrates her own considerable learning in both music theory and
theology---not to mention the elegance of her poetic craft---and also reveals
how much she is expecting her readers to know.%
    \Autocite{Stevenson:SorJuanaMusicalRapports}
Our greater knowledge of Sor Juana's life as compared to other villancico poets
enables us to recognize a degree of subjective identification in her image of
Mary as a \term{maestra de capilla}, since in Sor Juana's experience the only
female chapelmasters were nuns like herself; and we may question whether Sor
Juana envisions Mary as going beyond the earthly domains of female leadership
to direct a mixed chorus of the saints and the genderless angels. %XXX
% XXX relevant sor juan lit

As the succeeding chapters will show, when poetry like this was set to music,
composers had the opportunity to match this intricate musical-theological
discourse with another layer of symbols in the sounding music.
This conjunction of verbal and musical play on musical concepts was no
accident: the texts of villancicos were written specifically as lyrics for
musical compositions, as Juan Díaz Rengifo stated in one of the earliest
literary descriptions of the genre, and composers had every reason to favor
poems that gave them opportunities for clever musical crafstmanship.%
    \Autocite{Rengifo:ArteMetrica}
Juan Gutiérrez de Padilla (\cref{ch:padilla-voces}) and Joan Cererols
(\cref{ch:cererols-suspended}) both took finely wrought, Gongoresque texts with
musical conceits and added a rich musical commentary on those conceits in their
intricate settings.
Each chapter in \cref{part:unhearable-music} traces a family of related
villancicos with the same or similar texts and demonstrates that this type of
high-concept metamusical villancico served a special purpose for Spanish
musicians, enabling them to situate themselves within a tradition of
composition.
%}}}3

%{{{3 heavenly, angelic music
\subsubsection{Pointing to a Higher Music: Heavenly and Angelic Music}

Thus far we have seen how Spanish composers represented other kinds of music
like birdsongs, instrumental music, and dances of different social groups
within the villancico genre; and how they created songs that pointed to
themselves, whether by parodying the genre's own conventions or by using
solmization to draw listeners' attention to the act of singing itself.
How, then, did composers use villancicos to refer not to any kind of earthly
music, but rather to point to celestial (planetary) and heavenly (angelic,
divine) music?
When a villancico referred to the music of the spheres or to angelic music, the
music signified was impossible to hear with earthly ears, so the human music
would only function as a sign to the extent that a listener believed it to
correspond to what those higher forms of music sounded like, or understood them
to be mere imitations of something higher.
As we will trace in \cref{part:unhearable-music}, Spanish composers developed a
family of conventional tropes for evoking heavenly music.
One of these conventions was to set up a contrast between stylistic allusions
to distinct types of human music with different values in a hierarchy of
musical styles.
The most elevated form of earthly music, learned counterpoint in the by-then
classic style of Palestrina, was typically contrasted with more worldly types
of music, such as the rhythms of dance and the melody-and-accompaniment style
of popular songs.
This hierarchy of human musics was mapped on to the greater hierarchy of
earthly and heavenly music, so that old-style counterpoint stood in for
angelic and divine music, though really it was the contrast between musical
topics that gave it this meaning.
We have already seen how Juan Gutiérrez de Padilla contrasted the music of
Angolans and angels in his 1652 ethnic villancico, with the angels quoting
plainchant and singing in perfect species counterpoint.
This approach was used across confessional lines in early modern Europe,
with well-known Lutheran examples by Heinrich Schütz, Dieterich Buxtehude, and
J. S. Bach (all in music envisioning heavenly bliss after death).%
\begin{Footnote}
    \Autocites{Johnston:RhetoricalPersonification}
    {Yearsley:Buxtehude}{Yearsley:BachThron}.
    Lutherans saw the role of the boys' choir in relation to the congregation
    as analogous to that of the angelic chorus in relation to all human
    singers; see \autocite[\XXX]{Cashner:Gerhardt}.
\end{Footnote}

A typical example of the angelic trope is \wtitle{Angélicos coros con gozo
cantad}, a Christmas villancico by Antonio de Salazar that was part of the
repertoire of the Conceptionist sisters of the Convento de la Santísima
Trinidad in Puebla de los Ángeles (\cref{mux:Salazar-Angelicos_coros-1}).%
\begin{Footnote} 
    \sig{MEX-Mcen}{CSG.256}; edited in \autocite{Cashner:WLSCM32}.  
\end{Footnote}
Salazar (\circa{1650}--1715) was probably born in Puebla and may have sung in
the Puebla Cathedral chapel under Gutiérrez de Padilla; he served as
chapelmaster of Mexico City Cathedral from 1679.%
    \Autocite{Koegel:Salazar} % + DMEH
The convent collection features numerous pieces by Salazar, possibly composed
or arranged specifically for this community.% 
    \citXXX[catalog]
Puebla, the original American \quoted{city of angels}, was built on a site
believed to have been revealed by angels to the bishop of Tlaxcala, and
buildings and artworks dedicated to the angels, especially Saint Michael the
Archangel, are everywhere in the city.

%{{{5 music Salazar Angelicos coros 1
\insertMusic{Salazar-Angelicos_coros-1}
{Salazar, \wtitle{Angélicos coros con gozo cantad} (\sig{MEX-Mcen}{CSG.256}),
opening}
%}}}5

The anonymous poem echoes the first Responsory of Christmas Matins
(\quoted{Gaudet exercitus Angelorum}) as it invites the choirs of Christmas
angels to sing their \quoted{Gloria} over the stable in Bethlehem on the night
of Christ's birth.
Since \mentioned{Bethlehem} in Hebrew means \quoted{House of Bread}, the
villancico also celebrates the sacramental presence of Christ in the
Eucharistic host on the Christmas of Salazar's \soCalled{present day}.
Though the words speak to the angels, the musicians who sing these words also
play the part of the angels, so that hearers are invited to listen for the
angelic voices \emph{through} the voices of the church ensemble. 
The invocation to the angels is sung first by the Tiple I, in a gesture
beginning with a rising fifth and then falling by step, as though looking up to
the heavens and then following the angels' descent.
In the Puebla convent choir, this part was performed by \quoted{Madre Andrea},
whose name is written into her part.
As though answering the call, the other two voice parts of Chorus I enter in
\measures{2}, Tiple II in canonic imitation, and Alto I harmonizing with it
homorhythmically. 
In \measures{4-5} the second chorus joins with a similar imitative
pattern, until all join together in a lilting, dancelike cadence on
\foreign{cantad}.
Salazar uses contrapuntal imitation again on \foreign{celestes esquadras},
inverting the opening motive (\measures{14-22}).
For the command \foreign{bajad} (come down), Salazar switches from CZ
triple meter to duple (C or \term{compasillo}), and creates a cascading
contrapuntal passage passed from voice to voice, moving from high F\octave{5}
down to C\octave{3} (\cref{mux:Salazar-Angelicos_coros-2}).
The general affect of the piece seems gentle and sweet, partly because of the
largely static diatonic harmony and the lilting or dotted rhythms.

%TODO fix bar numbers

%{{{5 music Salazar Angelicos coros 2
\insertMusic{Salazar-Angelicos_coros-2}
{Salazar, \wtitle{Angélicos coros con gozo cantad} (\sig{MEX-Mcen}{CSG.256}):
Angels descending in imitative counterpoint}
%}}}5

All of these musical characteristics are typical ways that villancico composers
represented angelic music: especially contrapuntal imitation, in a reference to
the ordered music of heaven, and symbolic patterns of ascent and descent.
Salazar uses different styles of earthly music---particularly the contrast
between contrapuntal and homophonic styles---to point to the contrast between
different levels of music on a cosmic scale, between human music, music of the
spheres, and angelic song.
Because the triple-meter style of the first section, which asks the angel
choirs to sing, is more typical of villancico style, this part might be heard
to represent the actual singing of the angels.
The duple-meter section on \quoted{bajad} might be understood as a more literal
portrayal of the angels themselves.

The piece connects Boethian \term{musica instrumentalis} to the higher forms of
human and cosmic music. 
\soCalled{Heavenly} villancicos map a lower level of music onto a higher one
within the Neoplatonic cosmos, in which the perceptible \soCalled{world of
change and decay} is an imperfect reflection of a higher world of ideal forms.
Thus earthly music of any kind, metamusical or not, would always point beyond
itself to higher forms of music and ultimately to God.
Metamusical pieces intensify this aspect of music by calling the listener's
direct attention to the artifice of the music itself.
Such pieces give listeners the opportunity to rise in Neoplatonic contemplation
from what is heard by the ears to the higher music (ultimately of the divine
nature) that can only be discerned by the soul through faith.

This contemplative or meditative approach to music was favored among the
Spanish elite.
One venue in which this kind of listening was favored was in meetings or
sponsored services of religious confraternities.
Salazar belonged to the Confraternity of Saint Michael the Archangel.
A collection of printed sermons by the Carmelite preacher Fray Andrés de San
Miguel of Puebla includes a sermon Fray Andrés was invited to preach at a
gathering of Salazar's confraternity.
The title of the sermon explicitly mentions Salazar as having commissioned the
sermon, and this suggests that Salazar may have been head of the confraternity
as well. 
In the opening of the sermon, the preacher mentions that they are gathered in
the \quoted{church of the Encarnación}---possibly the Augustinian monastery in
Puebla.
The monk's self-deprecating introduction makes it clear that he is addressing
an elite congregation of highly learned and accomplished men.
The cleric expresses his concern that he does not really know enough music to
be addressing such a group, which suggests these were men specifically educated
in music.
%  \X cite source
The preacher's praise for Salazar (though expected if Salazar was responsible
for the paid commission) indicates a high level of respect and appreciation for
this chapelmaster.
He goes on to imply that there will be a musical performance at the same
liturgy, in which the audience will be able to hear this for themselves.
In fact, Fray Andrés says that Don Salazar could compose a better sermon in
music than he himself could preach in words. 

The sermon that follows is an exposition of the identity and deeds of
Saint Michael.
The friar structures this discussion not according to the verses or portions of
a Scriptural citation (as was more common), but according to each syllable of
the Guidonian hexachord.
Since Michael's name in Hebrew means \quoted{he who is like God}, or
\quoted{Quis ut Deus} in Latin, the friar uses UT to discuss Michael's name.
Since Michael is the chief of all the angels, he is their \quoted{king} or RE;
and so on.
As the preacher had warned, his musical knowledge does seem to have been rather
thin, since he does not use any other musical terminology or musical metaphors
in the sermon. 
The sermon is about angels, not about music; but it uses the terminology of
music as a framework to discuss the angels.

We do not know what music may have been performed at this service (or even what
kind of liturgy it was), but Salazar's \wtitle{Angélicos coros} would seem to
be the type of music that might have been chosen.
The version surviving in the Colección Jesús Sánchez Garza is arranged for the
sisters of the Convento de la Santísima Trinidad in Puebla---another
semi-private music venue with elite, high-level musical performance.
This is not Salazar's most learned composition, or even his most metamusical,
but it does allow us to imagine how themes of music were treated in villancicos
performed in closed and private spaces for circles of musical and theological
connoisseurs.
Villancicos about angels have a strong Neoplatonic charge to them: the singers
turn their attention heavenwards to address the angels directly (since angels
were believed to be present at every liturgy, \bibleverse{ICor}(11:10)), while
they also stand in for, or sing along with, their heavenly counterparts.
Angel pieces exhort the audience to lift their ears upwards as well, and ascend
in the chain of contemplation beyond even \emph{musica mundana} to the music of
the \emph{cielo Empyreo} in the highest Heaven.
%}}}3
%}}}2
%}}}1

%{{{1 theology of music
\section{Theological Listening in the Neoplatonic Tradition}

%{{{2 intro sources luis, kircher, augustine, boethius

% XXX apologize for long quotations? or cut them?
Villancicos on the subject of music consistently manifest a Neoplatonic
theological worldview, an understanding of which is necessary to grasp the
genre's religious functions.
Having drawn out aspects of this theology inductively from the examples of
metamusical compositions, in the final section of this chapter we may turn to
theological literature to establish a more systematic foundation.
The sixteenth and seventeenth centuries in Spain brought a revival of interest
in Neoplatonic theology in the tradition of Augustine, partly because of new
printed editions of his works (like the complete works set owned by the
Oratorian Society in Puebla, of which Juan Gutiérrez de Padilla was a member).%
    \citXXX[revivals, Oratorians]

Christian Neoplatonists followed Augustine in viewing the material
world as a reflection of a higher spiritual reality which ultimately had its
source in the Supreme Good which was the Godhead.%
\begin{Footnote}
    An important later source for this concept is the \wtitle{Spiritual
    Hierarchy} attributed to Dionysius the Areopagite.
\end{Footnote}
The material world reflected higher truths only imperfectly, but nevertheless
this world was also the only means through which those truths could be reached.
In connection with Catholic sacramental theology, material objects and physical
actions became means through which humans could encounter divine grace.
Neoplatonic contemplation could be understood as a dialectical process of
discerning the degree both of similarity and of dissimilarity between earthly
objects and heavenly truth.

Neoplatonic thought was promulgated in Spain through the copious writings of
the Dominican friar Luis de Granada (known as Fray Luis in Spanish), one of the
most widely read authors in the empire.%
    \citXXX[readership]
His work is a self-acknowledged synthesis of patristic and Classical
sources as well as a summary of common beliefs of his own time.
The 1589 \wtitle{Introducción del Símbolo de la Fe} (Introduction to the Creed)
presents a theological interpretation of the created world within this
tradition.
Fray Luis teaches that the natural world is a reflection of a higher
truth---God's own nature---and that the creation was given so that by
reflecting on it people would come to know its Creator.
In the friar's theology, music embodies the harmonious order of creation; he
suggests music could provide a way to contemplate creation and its
Creator.

Writers on music, too, developed a Neoplatonic approach, building on the
concept of the medieval philosopher Boethius that there were three kinds of
music (\cref{tab:Neoplatonic-hierarchy-music}).
At the lowest level is \term{musica instrumentalis}---music played and sounded,
music that humans can hear.
Higher up is \term{musica humana}---the harmony of body and soul, and of one
human being with another in society.
Still higher is \term{musica mundana}---the harmonies created by the perpetual
movement of the planetary spheres.
The three Boethian types of music are arranged hierarchically and each one
points beyond itself to a higher level.
The treatises used to teach musical composition in seventeenth-century Spain,
most notably Pedro Cerone's \wtitle{El melopeo y maestro} (1613) and Andrés
Lorente's \wtitle{El porqué de la música} (1672), present music within this
cosmology of music.

A key source for Neoplatonic thought on music is the encyclopedic
\wtitle{Musurgia universalis} of 1650 by Athanasius Kircher.%
    \Autocite{Kircher:Musurgia}
The Jesuit polymath's work was disseminated through Jesuit networks across the
globe: a copy was sent as far as Manila, and two copies are preserved today in
Puebla.%
\begin{Footnote}
    \Autocites{Findlen:Kircher}{Godwin:KircherTheater};
    on the worldwide distribution of the book as far as Manila, see
    \autocite[48--50]{Irving:Colonial}.  
    The book may be found in historical collections in Madrid, Barcelona, Mexico
    City, and Puebla (two copies). % XXX sigla
\end{Footnote}
Kircher describes in detail the latest scientific knowledge about the anatomy
of hearing and vocal production and the physiology of bodily humors and
affects; and lays out specific examples of how particular musical structures
work through these bodily systems.
Kircher presents a cosmic view of music according to Neoplatonic traditions of
theology and music theory, in which the whole universe is encompassed in the
\quoted{working of music}---a rough translation of his inventive
Greek-and-Latin title.%
        \citXXX[secondary Kircher lit]

%{{{5 table Neoplatonic hierarchy
\insertTable{Neoplatonic-hierarchy-music}
{Hierarchy of types of music in Neoplatonic thought, after Boethius (read from
bottom up)}
%}}}5
%}}}2

%{{{2 hearing book of nature
\subsection{Hearing the Book of Nature Read Aloud}

Fray Luis de Granada begins his \wtitle{Introduction to the Creed} with an
epitome of Neoplatonic-Augustinian thought: \quoted{The ultimate and highest
good of man}, he writes, \quoted{consists in the exercise and use of the most
excellent work of man, which is the knowledge and contemplation of God}.%
    \Autocite[182]{LuisdeGranada:Simbolo}
The created world, he teaches, is a \quoted{book of nature} in which is written
the grandeur, love, wisdom, and faithfulness of its Creator.
The first goal of humankind, then, is to learn to read this \quoted{book of
nature} in order to come through it to the knowledge of God. 
The goal of contemplating creation is \quoted{ascending by the staircase of the
creatures to the contemplation of the wisdom and beauty of the Maker}.%
    \Autocite[184]{LuisdeGranada:Simbolo}

The reason one can \quoted{read} God through nature, Fray Luis teaches, is that
the created world is a reflection of God's perfect order---a concept the friar
repeatedly expresses using musical metaphors.
Fray Luis compares the perfect order of nature to a harmonious musical
composition in which everything fits together \foreign{con sumo concierto}
(with the most perfect concord).
All the created things in this world, Fray Luis writes, \quoted{like concerted
music for diverse voices, harmonize together \addorig{concuerdan} in the
service of man, for whom they were created}.%
    \Autocite[191]{LuisdeGranada:Simbolo}
The movement of the heavenly spheres, and their effects on the earth, are like
a great \quoted{chain, or, it can be said, this dance, so well ordered, of the
creatures, and like music for diverse voices \Dots.
Because things so diverse could not be reduced to a single end with a single
order, if there were not one who was like a chapelmaster \addorig{maestro de
capilla}, who reduces them to this unity and consonance}.%
    \Autocite[191]{LuisdeGranada:Simbolo}

Everyone who compares something to music has some kind of earthly music in
mind.
When Fray Luis compares the world to music \quoted{in diverse voices} he
obviously has in his \quoted{mind's ear} polyphonic music of his own time, such
as he would have heard at the Portuguese Royal Chapel as confessor to the
queen.%
    \citXXX[names of composers]
Likewise, when he compares God to a \foreign{maestro de capilla}, that has all
the implications of that office in the Iberian context, which included
composition, teaching, and leading the choir in some form of conducting.%
\begin{Footnote}
    Compare the conceit of Christ as chapelmaster in \cref{ch:padilla-voces}).
\end{Footnote}
According to Fray Luis's metaphor, then, God is creator, prime mover, and
sovereign ruler over creation, actively and intimately involved in its ongoing
progress.

At the same time, these references to music are more than metaphors: the
universe for Neoplatonists is not only like music, it actually is musical in
its structure.
For Fray Luis, not only does creation reflect God's order; it actively
proclaims that fact.
It speaks or sings with its own voice to communicate God's glory to the human
who knows how to listen.
Paraphrasing Augustine's preaching, the friar writes: \quoted{Look around at
all these many things from the heaven to the earth, and you will see that they
all sing and preach their Creator; because all types of creatures are voices
\add{or, utterances} that sing his praises}.% 
    \Autocite[185, glossing Augustine's commentary on Psalm 26]
    {LuisdeGranada:Simbolo} 
While the full knowledge of God can only come with the aid of divine revelation
through the Scriptures and the Church, Fray Luis praises God that humans can
study his nature in \quoted{the university of created things, which declare to
us \add{literally, \quoted{give us voices}} that you love us, and teach us why
we should love you}.% 
    \Autocite[186]{LuisdeGranada:Simbolo}
Fray Luis acknowledges, however, that apart from angels and birds, most of
creation is mute and does not literally have its own voice with which to
communicate its message of divine glory.
This \quoted{message} is not a linguistic one, but rather, their message is
simply themselves: in the created world, the medium is the message.
\quoted{Now these admirable works do not speak or testify this with human
voices \Dots}, Fray Luis writes, \quoted{rather their speech and testimony is
their invariable order and their beauty, and the artifice with which they are
so perfectly made, as though they were made with a ruler and plumb line}.%
    \Autocite[192]{LuisdeGranada:Simbolo}

In this theological system, music has unique value because it actually provides
a voice through which creation can make audible its message-of-being.
If nature was a book, for a thinker in 1580s Spain, then it was meant to be
read out loud.
As Margit Frenk has documented, books in this period were not read silently,
but required someone to give them voice, and tomes like Fray Luis's devotional
books were written with that intention.%
    \Autocite{Frenk:Voz}
One edition of Fray Luis's own book \wtitle{Doctrina Cristiana} was published
with a notice at the beginngin from the Archibishop of Toledo granting a
certain number of days of indulgence for each paragraph that anyone
\quoted{read or heard read} (that is, had read to them).%
    \citXXX[signature]
To read the \quoted{book of nature}, therefore, someone must perform it
vocally---and this is what music could do.
In the Christian Neoplatonic tradition, human music unlocks the musical voice
contained within the substance of created things.
Through metal pipes, horns, and bells; through wood viol cases, gut strings,
and skin drums; even through reverberant stone church walls, the very matter of
creation is made to resound with the perfectly ordered mathematical-harmonic
proportions placed within it by the Creator---proportions which themselves
reflect God's own perfect order.
%}}}2

%{{{2 voice microcosm
\subsection{Voice as Expression of Man, the Microcosm}

If pipes and strings testify to the order of creation, then the human body as
the microcosm of creation is the ultimate instrument through which nature is
given voice.
Fray Luis concludes his exposition of the six days of creation (based largely
on the \emph{Hexameron} of Saint Basil) by saying that God's creation of man on
the sixth day was like the conclusion of an oration, when the speaker draws
together all his themes into a final epitome.
Thus man is the summation of all that God had created in the previous five days
and encompasses them all within himself.% 
    \Autocite[243]{LuisdeGranada:Simbolo}

When Athanasius Kircher (in the tenth book of the \emph{Musurgia}) continues
this hexameral tradition with his own treatment of the six days of creation, he
replaces the rhetorical metaphor with a musical one.
Instead of creation being God's oration, Kircher presents it as a musical
improvisation (a \quoted{Praeludium}) on God's cosmic organ (see
\cref{ch:cererols-suspended}).%
        \Autocite[\range{vol}{2}, 366--367]{Kircher:Musurgia}
On the sixth day, Kircher says, God recapitulates all his themes and pulls out
all the stops by creating man.
Here again, the comparison to music must be based on some actual music known to
Kircher; his description closely resembles the structure of a Praeludium by the
likes of Dieterich Buxtehude, which develops a motivic kernel through various
sections and culminates in a fugue for the full organ.%
    \citXXX[check dates, cite rhetorical analysis of buxtehude]
In Kircher's worldview, all the systems and elements of creation (stars,
planets, humors, rocks, animals, and so on) intersect in the individual human
body.% 
    \Autocite[vol. 2, 402]{Kircher:Musurgia}

For Kircher, the human voice is the unique expression of the individual,
reflecting each person's unique temperament and blend of the four humors.%
    \Autocite[\range{vol}{1}, 23--24]{Kircher:Musurgia}
Kircher defines the voice thus: \quoted{The voice is a living sound \add{or,
sound of the soul}, produced by the glottis through the percussion of respired
breaths that serve to express the affects of the soul}.% 
    \Autocite[\range{vol}{1}, 20]{Kircher:Musurgia}
Since each voice is unique, only in concert do voices fully reflect nature and
nature's God.
Cantus, Altus, Tenor, and Bass parts provide a place for all types of human
voices, Kircher explains, and correspond respectively to fire, air, water, and
earth.
Thus they form a choral microcosm both of humanity and of all creation.%
        \Autocite[\range{vol}{1}, 217--219]{Kircher:Musurgia}

Fray Luis also ventures an explanation of the human voice in both musical and
theological terms.
He exalts the voice as the audible expression of the human body and
vocal music as the most perfect kind of music.%
    \Autocite[243]{LuisdeGranada:Simbolo} 
Fray Luis praises the human voice as the highest of all musical instruments
(indeed, as the paradigm for them), as a means of forming social relationships
between people, and as a form of communication between human and divine:
\begin{quoting} 
    The lungs also serve to create the voice, because, when the air that they
    exhale leaves them with a great impetus, and touches the voicebox or
    \quoted{little bell} that we have at the entrance of the lungs, the voice
    is formed. \Dots{}
    But here it is to be noted that the mouth of the pipe coming out of the
    lungs is neither large nor round, but is drawn tight \add{hendida} just like
    the slot of an alms box,
    which opening serves to form the voice; this is why the mouths of flutes
    and dulcians are constructed in this fashion, because in this manner, the
    compressed air entering through them, the voice is caused.%
    \Autocite[252]{LuisdeGranada:Simbolo} 
\end{quoting}
Flutes and dulcians would be the example ready at hand for the friar because
these instruments were commonly played in the Iberian church music he knew.
But unlike a wooden flute, he says, the voice can take on any shape needed and
is as unique as each person's body.
The voice therefore expresses human individuality, and voices of different
types in concert enact harmony between people: 
\begin{quoting} 
    Moreover, here is a thing worthy of much consideration, to see the
    omnipotence and wisdom of the Creator, who was able to form something like
    a flute from flesh, which serves for singing.
    For to make a flute or trumpet from a solid material such as wood or some
    metal, is not much; for the hardness of the material serves for the
    resonance of the voice.  
    But to  make this out of flesh (such as is the windpipe of the lungs), and
    such that through it some voices are formed of women and of men, so sweet
    that they seem more like those of angels than of humans, and these with
    such variety of notes \add{punctos}, without having the finger holes of
    flutes that provide this variety, this is something that declares the power
    and the wisdom of that sovereign artisan, who in such a manner forged the
    flesh of this windpipe so that in it could be formed a voice sweeter and
    milder than that of all the flutes and instruments that human industry has
    invented.

    And there is no end of admiration for the variety that there is in this for
    the service of harmonious music \add{música acordada}.  
    For some throats are narrow, in which are formed the trebles \add{tiples},
    and others in which are formed voices so full and resonant that they seem
    to thunder through an entire church, without which there could not be
    perfect music.

    All of which that divine presider traced and ordained, so that with this
    mildness and melody the divine offices and their praises should be
    celebrated, with which to awaken the devotion of the faithful.%
        \Autocite[252]{LuisdeGranada:Simbolo} 
\end{quoting}

Fray Luis wants his readers to hear God's glory reflected most fully in the
concerted harmony of diverse human voices, which he says were created for the
purpose of singing of singing in divine worship.
The voice in church is the definitive example of vocal music for Fray Luis.
Sacred polyphony glorifies God, then, simply by realizing the potential for
which the voice (and the body) was made.
In the above passage, the sound of the voice alone proclaims God's power and
wisdom just in itself, apart from whatever words or musical figures it might
articulate.

Fray Luis sees speech as something \quoted{added} to the voice, which makes it
possible for the voice to communicate and form social relationships:
\begin{quoting} 
    Now here it is to be noted that when to the voice which proceeds from this
    place is added the instrument of the tongue, we come to articulate and make
    distinctions with this voice, and thus is formed speech, serving us by this
    instrument and punctuating \add{hiriendo} with it sometimes in the teeth
    and other times in the interior of the mouth.
    And just as the flute produces different sounds by touching on different
    holes, likewise the tongue, touching in different parts of the mouth, forms
    different words.  
    By this manner the Creator gave us the faculty to speak and communicate our
    thoughts and concepts to other men.% 
        \Autocite[252]{LuisdeGranada:Simbolo}
\end{quoting}
Fray Luis might see music---with its own system of articulations and
distinctions---as another way to \quoted{communicate our thoughts and concepts}
just as well as spoken language, but he also presents music as a product of the
voice before any articulation is added.
This definition of voice would mean that in vocal music there are always two
layers---the articulated \quoted{speech} aspects, and behind these the wordless
sustained voice.
Citing Augustine's \wtitle{De doctrina christiana} (the classic exposition of
Christian preaching and teaching), Fray Luis---who was himself the author of
six volumes about \wtitle{Rhetorica ecclesiastica}---says that the main task of
the student of rhetoric is to hear and identify the rhetorical tropes and
techniques used by another orator.
In the same way, he says, the first task of humankind is to be a student of the
natural world, and to learn to recognize in creation the signs of God's
artifice as the Creator, which manifest his glory.

This would mean that the hearer of music could and should seek out this level
of musical structure while listening.
In a polyphonic vocal piece like a villancico, the bulk of musical structure is
borne by the sustained tones of the voice, singing vowels.
Apart from the words being sung, musical elements like mode, meter, motivic
development, and stylistic or topical allusions are all communicated by these
musical voices, and not simply by the voice as the bearer of words.
Music could thus reflect the divine through its sonic structure, apart from any
sacred linguistic meaning that may be attached as well.
If music's value and sacredness are not comprised solely in the words being
sung, then one must know how to hear the musical structure in order to receive
the full benefit.

Listening to music within a Neoplatonic worldview, according to these
theological sources, may be summarized as follows:
\begin{enumerate}
\item Music is a reflection of the natural order.
\item The natural order is itself a reflection of God.
\item By paying attention to nature one can come to know and believe in its
    Maker.
\item Therefore listening to music may be a primary way of \quoted{reading the
    book of nature} and coming to faith in nature's Creator.
\item Music, especially vocal music, conveys sacred meaning to those who know
    how to listen, even apart from words.
\item The performance of music actively creates concord in society and between
    people and God.  
\end{enumerate}

Metamusical villancicos explicitly emphasize the challenge that was central to
all music-making in the Christian Neoplatonic tradition, to use the imperfect
medium of sounding music to evoke all the higher forms of music, to lead
listeners in contemplation up the chain of being beyond simply what was heard.
The recurrence of the \quoted{Listen!} exordium in villancicos may indicate
that the genre itself was fundamentally about getting people to listen.
Metamusical pieces drew listeners' attention to the artifice of the music they
were hearing and pointed them toward a higher form of music.
These villancicos invited listeners not simply to hear, but to \quoted{take
heed}, to both discern deeper meanings in what they heard and to
put what they hear into practice.
Understanding how Spanish Catholics these links between hearing and faith, and
the challenges they faced in using music to connect one to the other, will be
the challenge of the next chapter.
%}}}2
%}}}1

\endinput
