% Sources of Suspended, cielos
\begin{tabulary}{\linewidth}{rr Lp{5em} LLc}
    \toprule
    No.
    & Year
    & City, Institution
    & Feast 
    & Composer (Poet)
    & Source
    & Family \\

    \midrule
    \multicolumn{7}{c}{\emph{Poetry Imprints}} \\

    \addlinespace
    1 
    & 1651 
    & Madrid, Royal Chapel 
    & Christmas 
    & Carlos Patiño? 
    & \sig{E-Mn}{R/34199/27}
    & A \\

    \addlinespace
    2
    & 1668
    & Calatayud, Colegiata Sta. María la Mayor 
    & Christmas 
    & Juan Muñoz 
    & \sig{GB-Lbl}{11450.dd.8 (54)}
    & A \\

    \addlinespace
    3
    & 1675
    & Alcalá, Iglesia SS. Justo y Pastor
    & Christmas 
    & Antonio Garcia? (Attr. Manuel de León Marchante)
    & León Marchante, \wtitle{Obras poéticas} (Madrid, 1733), 139 
    & B \\

    \addlinespace
    4
    & 1680
    & Seville, Cathedral
    & Christmas 
    & Alonso Xuárez?
    & \sig{E-Mn}{VE/83/10}
    & B \\

    \addlinespace
    5
    & 1681
    & Seville, Iglesia S. Salvador
    & Christmas 
    & Miguel Mateo de Dallo y Lana 
    & \sig{E-Mn}{VE/79/7}
    & B \\

    \addlinespace
    6 
    & 1683
    & Zaragoza, Basilica El Pilar
    & Epiphany 
    & Diego de Cáseda
    & \sig{E-Mn}{VE/129/2}, \sig{GB-Lbl}{1073.k.22(07)}
    & B \\

    \addlinespace
    7
    & 1689
    & Madrid, Royal Chapel
    & Christmas 
    & Juan de Navas?
    & \sig{E-Mn}{VE/88/80}
    & A+B \\

    \midrule
    \multicolumn{7}{c}{\emph{Music Manuscripts}} \\

    \addlinespace
    8
    & \circa{1660}
    & Canet de Mar, Església de Sant Pere i Sant Pau (orig., Abbey of Montserrat?) 
    & Christmas (mislabeled Eucharist)
    & Joan Cererols 
    & \sig{E-CAN}{AU/116}
    & A \\

    \addlinespace
    9
    & \circa{1680}
    & Barcelona, Unknown (Same origin as S8)
    & Eucharist
    & Anon. variant of Cererols
    & \sig{E-Bbc}{M/765/25}
    & A  \\

    \addlinespace
    10
    & \circa{1690}
    & Ibarra, Ecuador, Conceptionist Convent (from Quito Cathedral?) 
    & Christmas, Sanctoral, General
    & Anon. (Quito chapelmaster?) 
    & Ecuador, Ibarra, Parish archive 
    & B′ \\ % prime U+2032

    \bottomrule
\end{tabulary}
\endinput
